---
title: "Kickstarter"
author: "Ang Ming Kwan"
date: "May 23, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1. Installing packages

For the purpose of this exercise, 3 R packages will be used. They are tidyverse, jsonlite and SmartEDM.

In the code chunk below, p_load() of pacman package is used to load the R packages into R environment.

```{r}
pacman::p_load(tidyverse, jsonlite,
               tidygraph, ggraph)
```

## Importing graph data

For the purpose of this exercise *mc2.json* file will be used. Before getting started, you should have the data set in the *data* folder. We will also extract the data from the *list* into data frames and create *rds* file formats of the data frames to allow faster extractions for subsequent uses.

```{r}
mc1_data <- fromJSON("data/MC1_final_00.json")
```

```{r}
mc2_data <- fromJSON("data/MC2 data.json")
```

```{r}
write_rds(mc1_data, "data/mc1_data.rds")
```

```{r}
org_chart <- fromJSON("data/org_chart.json")
```

```{r}
mc2_df <- as.data.frame(mc2_data$events)
```

```{r}
write_rds(mc2_df, "data/mc2_df.rds")
```
