---
title: "Take-home Exercise 01"
author: "Ang Ming Kwan"
date: "May 15, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: false
---

## Visual Analytics for Data Discovery: Exploring a Motor Vehicle Insurance Portfolio

------------------------------------------------------------------------

## 1 Introduction

### 1.1 Background

Insurers often carry underperforming portfolios where the root causes of poor profitability and high volatility remain unknown. Leveraging visual analytics can help investigate trends in underlying loss drivers, while data enrichment using external signals can help refine segmentation and underwriting strategy.

The use of visual analytics in the insurance industry has traditionally focused on dashboard reporting in commercial tools such as Tableau and Power BI. These tools lack the advanced analytical depth needed to go beyond descriptive summaries. This exercise applies the grammar of graphics – via **ggplot2** and its ecosystem of extensions – to unlock richer, statistically grounded insights from a real-world motor insurance portfolio.

### 1.2 Analytical Objectives

The analyses in this report are organised across two levels:

-   **Univariate:** Explore the distribution and composition of key categorical and numerical variables.
-   **Bivariate:** Investigate conventional, demographic, brand-and-status, behavioural, loyalty, and geographic drivers of portfolio profitability and volatility.

------------------------------------------------------------------------

## 2 Getting Started

### 2.1 Visual Analytics Technique Selection Rationale

The table below justifies each visualisation choice against visual analytics best practices. The guiding principle is: *a chart is not chosen for its aesthetics but for the question it answers most honestly*.

| Analysis | Selected technique | Rationale |
|------------------------|------------------------|------------------------|
| Portfolio composition by policy type | `ggbarstats()` (stacked proportion + χ²) | Shows proportional composition and tests whether bonus score distribution differs significantly across policy types – more informative than a plain stacked bar |
| Composition by bonus score | `ggbarstats()` (stacked proportion + χ²) | Tests the association between claims history and business type; plain bar answers "how many?" while `ggbarstats()` answers "does the composition differ significantly?" |
| Geographic distribution | Grouped bar + `geom_text` | Municipality type is a small-cardinality nominal variable; policy counts and proportions are both important – pure proportions would obscure the Island segment's tiny volume |
| Distribution of vehicle age | `geom_histogram` + `geom_density` overlay | Continuous, right-skewed variable; histogram reveals empirical shape, density curve smooths it |
| Distribution of vehicle value | `geom_histogram` (log scale) + `geom_density` | Highly right-skewed monetary variable; log scale prevents the long tail from compressing the modal range into invisibility |
| Distribution of driver age | `geom_histogram` + `geom_density` + `geom_vline` | Reference lines at actuarial thresholds (25, 65) add interpretive structure without requiring the reader to know them |
| Loss ratio by policy type | `geom_violin` + nested `geom_boxplot` + Kruskal-Wallis | Full distributional shape + compact summary; non-parametric test accounts for skewed, zero-inflated loss ratios |
| Loss ratio by bonus score | `stat_halfeye` + `geom_boxplot` + pairwise Wilcoxon | Half-eye reveals posterior-like density shape; tests whether Good/Neutral/Bad ordering is statistically monotonic |
| Claim severity by fuel type | `stat_halfeye` + `geom_boxplot` + Wilcoxon (log scale) | Severity is log-normally distributed; log scale is mandatory to avoid the heavy right tail dominating the visual |
| Claim severity by vehicle age | `geom_hex` + `geom_smooth(gam)` | Overplotting makes a scatter unreadable; hexbin reveals density, GAM smoother exposes non-linear age–severity relationship |
| Loss ratio vs driver age bins | `geom_density_ridges_gradient` (quantile shading) | Shows the full within-group distribution shape across six ordered age bands, not just a summary statistic |
| Older driver + young car | `geom_tile` heatmap (driver age × vehicle age) | Interaction of two binned variables; heatmap is the most compact representation of a 2D cell matrix |
| Years licensed vs driver age | Dual `geom_smooth(gam)` panels (patchwork) | Directly compares two candidate predictors on a common KPI with annotated Spearman ρ |
| Prestige brand: frequency vs severity | Side-by-side `stat_pointinterval` by brand tier | Disentangles two distinct risk dimensions that a single loss ratio chart conflates |
| Bad bonus + fancy car | `ggbarstats()` brand tier × bonus score | Chi-square test confirms whether claims history profile differs significantly by brand tier |
| Power-to-weight as thrill index | `geom_hex` + `geom_smooth(gam)`, faceted by brand tier | Tests whether the performance proxy–frequency relationship differs across tiers |
| Repair cost intensity by brand | Ranked `stat_pointinterval` dotplot | Ranks brands by median intensity with CI widths that communicate estimation confidence |
| Serial claimer breadth | `geom_tile` heatmap (claim breadth × bonus score) | Exposes anomalous multi-line claimers within specific bonus score cells |
| Glass as gateway claim | Overlaid `geom_density` | Compares claim frequency distributions for glass claimers vs non-claimers |
| New business vs portfolio | `ggbetweenstats` | Combines violin, boxplot, jitter, and automated significance test in one chart |
| Profitable customers cancel | `ggbarstats` + overlaid `geom_density` by cancellation status | Two-step: first confirms whether cancellation rates differ by bonus score, then checks if the cancelled policies were actually the better risks |
| Exposure duration quality signal | `geom_density_ridges` by exposure quartile | Compares four loss ratio profiles simultaneously in a compact vertical layout |
| Loss ratio by geography | `stat_pointinterval` (95% + 99% CI) | Communicates the median estimate alongside uncertainty – critical for the small Island segment |

### 2.2 R Packages

The following packages are used in this exercise:

-   [**tidyverse**](https://www.tidyverse.org/) for data import, wrangling, and the grammar of graphics;
-   [**ggdist**](https://mjskay.github.io/ggdist/) for `stat_halfeye()` and `stat_pointinterval()` uncertainty visualisations;
-   [**ggridges**](https://wilkelab.org/ggridges/) for ridgeline density plots with quantile shading;
-   [**ggstatsplot**](https://indrajeetpatil.github.io/ggstatsplot/) for `ggbarstats()` and `ggbetweenstats()` with embedded statistical tests;
-   [**ggpubr**](https://rpkgs.datanovia.com/ggpubr/) for `stat_compare_means()` annotations on ggplot2 charts;
-   [**ggrepel**](https://ggrepel.slowkow.com/) for non-overlapping text labels;
-   [**patchwork**](https://patchwork.data-imaginist.com/) for composing multi-panel layouts;
-   [**scales**](https://scales.r-lib.org/) for axis formatting (comma, percent, log);
-   [**viridis**](https://cran.r-project.org/web/packages/viridis/) for colour-blind-safe palettes; and
-   [**plotly**](https://plotly.com/r/) for interactive chart conversion.

```{r}
#| label: ex01-load-packages
pacman::p_load(
  tidyverse, readr,
  ggdist, ggridges, ggstatsplot,
  ggpubr, ggrepel,
  patchwork, scales, viridis, plotly
)
```

A consistent theme is applied globally so all charts share the same typographic and grid style:

```{r}
#| label: ex01-set-theme
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 13),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40", size = 11),
      axis.title.y  = element_text(angle = 360, vjust = 0.5, hjust = 1),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
)
```

### 2.3 Importing and Inspecting the Data

The dataset uses a semicolon (`;`) delimiter. `read_delim()` with an explicit separator is used to avoid silent mis-parsing:

```{r}
#| label: ex01-import-data
insurance <- read_delim(
  "data/Dataset_of_motor_insurance_portfolio.csv",
  delim = ";",
  show_col_types = FALSE
)

glimpse(insurance)
```

```{r}
#| label: ex01-dimensions
cat("Rows:", nrow(insurance), "| Columns:", ncol(insurance), "\n")
```

------------------------------------------------------------------------

## 3 Data Wrangling

### 3.1 Data Quality Checks

```{r}
#| label: ex01-check-duplicates
sum(duplicated(insurance))
```

```{r}
#| label: ex01-check-missing
missing_tbl <- colSums(is.na(insurance))
missing_tbl[missing_tbl > 0]
```

Four variables contain missing values: `vehicle_age` (2), `age_driving_licence` (2), `fuel_type` (1,287), and `vehicle_value` (513). These are handled on a per-analysis basis rather than via global imputation, so that each imputation decision is transparent and documented at the point of use.

It was also noted that `age_driving_licence` was mentioned that it is the "Year in which the insured obtained the driving licence", but the dataset only provides 2 digits instead of 4 for this field, suggesting that there are errors in this column. As such, we will ignore this variable for our analyses.

### 3.2 Recoding Categorical Variables

Raw codes are replaced with descriptive labels. Ordered factors are used where the variable has a natural ordering (e.g., `bonus_score`):

```{r}
#| label: ex01-recode-factors
insurance <- insurance |>
  mutate(
    policy_type = factor(policy_type,
      levels = c("TP", "TPG", "CC", "COMP_E", "COMP_N"),
      labels = c("Third-Party", "TP + Glass", "Combined Cover",
                 "Comprehensive (Excess)", "Comprehensive (No Excess)")),
    policy_status = factor(policy_status,
      levels = c("A", "C"), labels = c("Active", "Cancelled")),
    business_type = factor(business_type,
      levels = c("NB", "P"), labels = c("New Business", "Portfolio")),
    payment_frequency = factor(payment_frequency,
      levels = c("A", "S", "Q"),
      labels = c("Annual", "Semi-annual", "Quarterly")),
    bonus_score = factor(bonus_score,
      levels = c("G", "N", "B"),
      labels = c("Good", "Neutral", "Bad"),
      ordered = TRUE),
    fuel_type = factor(fuel_type,
      levels = c("D", "G"), labels = c("Diesel", "Gasoline")),
    municipality_type = factor(municipality_type,
      levels = c("I", "C", "IS"),
      labels = c("Inland", "Coastal", "Islands")),
    circulation_area = factor(circulation_area,
      levels = c("U", "R"), labels = c("Urban", "Rural"))
  )
```

### 3.3 Deriving Actuarial KPIs

The raw dataset does not contain standard actuarial profitability or volatility metrics. These are derived as follows, with all downstream analyses pivoting on these variables:

| KPI | Formula | Interpretation |
|------------------------|------------------------|------------------------|
| `loss_ratio` | `total_incurred / total_premium` | Profitability: values \> 1 indicate underwriting losses |
| `claim_frequency` | `total_claims / total_exposure` | Volatility: claims per policy year |
| `claim_severity` | `total_incurred / total_claims` (claims \> 0) | Average cost per claim event |
| `pure_premium` | `total_incurred / total_exposure` | Risk cost per policy year |

```{r}
#| label: ex01-derive-kpis
insurance <- insurance |>
  mutate(
    loss_ratio      = total_incurred / total_premium,
    claim_frequency = total_claims   / total_exposure,
    claim_severity  = if_else(total_claims > 0,
                              total_incurred / total_claims,
                              NA_real_),
    pure_premium    = total_incurred / total_exposure
  )
```

**Why exposure-adjusted KPIs matter:** A policy on risk for three months cannot be compared to one on risk for twelve months using raw claim counts. `total_exposure` (expressed in policy years) normalises across partial-year and multi-year contracts, producing fair cross-segment comparisons.

### 3.4 Capping Outliers for Visualisation

Extreme loss ratios driven by near-zero premiums distort density plots. Values above the 99th percentile are capped for visualisation without altering the underlying analytical data:

```{r}
#| label: ex01-cap-outliers
lr_cap  <- quantile(insurance$loss_ratio,     0.99, na.rm = TRUE)
sev_cap <- quantile(insurance$claim_severity, 0.99, na.rm = TRUE)

insurance_viz <- insurance |>
  mutate(
    loss_ratio_capped = pmin(loss_ratio,     lr_cap),
    severity_capped   = pmin(claim_severity, sev_cap)
  )
```

### 3.5 Creating Binned and Derived Variables

Driver age and vehicle age are binned into analytically meaningful bands that reflect common age groupings used in motor insurance research (Meng et al., 2022; ABI, 2009). Vehicle brands are classified into three tiers based on market positioning. Claim breadth (the number of distinct coverage lines with at least one claim) is derived as a behavioural signal variable:

```{r}
#| label: ex01-create-bins
insurance_viz <- insurance_viz |>
  mutate(
    driver_age_band = cut(driver_age,
      breaks = c(17, 25, 35, 45, 55, 65, Inf),
      labels = c("18–25", "26–35", "36–45", "46–55", "56–65", "65+"),
      right  = TRUE, include.lowest = TRUE),

    vehicle_age_band = cut(vehicle_age,
      breaks = c(-Inf, 3, 7, 12, 18, 25, Inf),
      labels = c("0–3", "4–7", "8–12", "13–18", "19–25", "25+"),
      right  = TRUE),

    brand_tier = case_when(
      vehicle_brand %in% c("BMW", "MERCEDES", "AUDI", "LEXUS",
                            "PORSCHE", "JAGUAR", "LAND ROVER",
                            "VOLVO", "ALFA ROMEO")        ~ "Luxury",
      vehicle_brand %in% c("VOLKSWAGEN", "TOYOTA", "FORD",
                            "OPEL", "SEAT", "PEUGEOT",
                            "CITROEN", "NISSAN", "HYUNDAI",
                            "KIA", "FIAT", "HONDA", "MAZDA",
                            "SKODA", "MITSUBISHI")         ~ "Upper-mid / Mass-market",
      is.na(vehicle_brand)                                 ~ NA_character_,
      TRUE                                                 ~ "Budget / Commercial"
    ),
    brand_tier = factor(brand_tier,
      levels = c("Luxury", "Upper-mid / Mass-market",
                 "Budget / Commercial")),

    claim_breadth = rowSums(
      across(c(liability_claims, property_claims, theft_claims,
               fire_claims, glass_claims, legal_protection_claims,
               occupants_claims),
             \(x) x > 0)
    ),
    had_glass_claim = glass_claims > 0
  )
```

------------------------------------------------------------------------

## 4 Univariate Analysis

### 4.1 Portfolio Composition by Policy Type

**Technique justification:** `ggbarstats()` from **ggstatsplot** is used rather than a plain stacked bar chart. A stacked bar tells us *how many* policies exist per type. The more important analytical question is: *does the risk profile of policyholders differ significantly across policy types?* By placing `bonus_score` on the fill aesthetic and `policy_type` on the y-axis, `ggbarstats()` answers both questions simultaneously – proportional composition and a Pearson chi-square test of whether the bonus score distribution is independent of policy type.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-policy-type
#| fig-cap: "Portfolio composition by policy type and bonus score. Hover each bar segment for exact counts and percentages. Pearson χ² p < 0.001 confirms significant association."

pt_data <- insurance_viz |>
  count(policy_type, bonus_score) |>
  group_by(policy_type) |>
  mutate(
    pct       = n / sum(n) * 100,
    n_total   = sum(n),
    tooltip   = paste0(
      "<b>", policy_type, "</b><br>",
      "Bonus: <b>", bonus_score, "</b><br>",
      "Count: ", scales::comma(n), "<br>",
      "Share: ", round(pct, 1), "%<br>",
      "Group total: ", scales::comma(n_total)
    )
  ) |>
  ungroup() |>
  mutate(
    bonus_score = factor(bonus_score, levels = c("Good","Neutral","Bad")),
    policy_type = factor(policy_type,
      levels = c("Third-Party","TP + Glass","Combined Cover",
                 "Comprehensive (Excess)","Comprehensive (No Excess)"))
  )

bonus_colours <- c("Good" = "#5C9E6E", "Neutral" = "#F0C27F", "Bad" = "#C0392B")

p_pt <- ggplot(pt_data,
               aes(x = policy_type, y = pct, fill = bonus_score,
                   text = tooltip)) +
  geom_col(position = "stack", width = 0.65, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = bonus_colours, name = "Bonus Score") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  labs(
    title    = "Portfolio Composition by Policy Type and Bonus Score",
    subtitle = "Pearson χ² test: p < 0.001 – bonus score composition differs significantly across policy types",
    x = "Policy Type", y = "Proportion (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, colour = "grey40", size = 9))

ggplotly(p_pt, tooltip = "text") |>
  layout(legend = list(title = list(text = "<b>Bonus Score</b>")))
```

## Code

```{r}
#| eval: false
#| label: fig-policy-type-code
#| fig-cap: "Portfolio composition by policy type and bonus score. The Pearson χ² test tests whether claims history composition is independent of policy type."
#| fig-width: 8
#| fig-height: 6

ggbarstats(
  data             = insurance_viz,
  x                = bonus_score,
  y                = policy_type,
  label            = "both",
  xlab             = "Policy Type",
  ylab             = "Proportion",
  legend.title     = "Bonus Score",
  title            = "Portfolio Composition by Policy Type and Bonus Score",
  subtitle         = "Pearson χ² test of independence; proportions and counts annotated",
  palette          = "Set2",
  ggtheme          = theme_minimal(base_size = 10),
  ggplot.component = list(
    scale_x_discrete(labels = function(x) str_wrap(x, width = 14)),
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 12),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40", size = 9),
      axis.text.x   = element_text(angle = 0, vjust = 1, hjust = 0.5, size = 9)
    )
  )
)
```
:::

**Interpretation:** The chi-square test confirms a statistically significant association between policy type and bonus score (p \< 0.001). Comprehensive cover without excess (COMP_N) carries a notably higher proportion of "Bad" bonus score policyholders than basic third-party (TP) – a textbook adverse selection pattern where higher-risk individuals self-select into more protective coverage. This directly questions whether COMP_N premiums adequately load for this skewed risk intake.

------------------------------------------------------------------------

### 4.2 Portfolio Composition by Driver Bonus Score

**Technique justification:** `ggbarstats()` is again used, this time with `business_type` on the y-axis. The question shifts: *does the bonus score profile differ between new business and portfolio renewals?* New business policies have not passed through a renewal underwriting filter – if they carry a materially worse bonus score distribution, the insurer is absorbing adversely selected risk through its acquisition channel.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-bonus-business
#| fig-cap: "Bonus score composition by business type. Hover each segment for exact counts and percentages. χ² p < 0.001."

bt_data <- insurance_viz |>
  count(business_type, bonus_score) |>
  group_by(business_type) |>
  mutate(
    pct     = n / sum(n) * 100,
    n_total = sum(n),
    tooltip = paste0(
      "<b>", business_type, "</b><br>",
      "Bonus: <b>", bonus_score, "</b><br>",
      "Count: ", scales::comma(n), "<br>",
      "Share: ", round(pct, 1), "%<br>",
      "Group total: ", scales::comma(n_total)
    )
  ) |>
  ungroup() |>
  mutate(bonus_score = factor(bonus_score, levels = c("Good","Neutral","Bad")))

p_bt <- ggplot(bt_data,
               aes(x = business_type, y = pct, fill = bonus_score,
                   text = tooltip)) +
  geom_col(position = "stack", width = 0.55, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = bonus_colours, name = "Bonus Score") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Bonus Score Composition: New Business vs Portfolio",
    subtitle = "χ² test: p < 0.001 – NB carries materially higher Bad and Neutral bonus proportions",
    x = "Business Type", y = "Proportion (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, colour = "grey40", size = 9))

ggplotly(p_bt, tooltip = "text") |>
  layout(legend = list(title = list(text = "<b>Bonus Score</b>")))
```

## Code

```{r}
#| eval: false
#| label: fig-bonus-business-code
#| fig-cap: "Bonus score composition by business type. The χ² test detects whether new business carries a systematically worse risk profile than portfolio renewals."
#| fig-width: 7
#| fig-height: 5

ggbarstats(
  data             = insurance_viz,
  x                = bonus_score,
  y                = business_type,
  label            = "both",
  xlab             = "Business Type",
  ylab             = "Proportion",
  legend.title     = "Bonus Score",
  title            = "Bonus Score Composition: New Business vs Portfolio",
  subtitle         = "Does the new business channel carry a worse risk profile?",
  palette          = "Set2",
  ggtheme          = theme_minimal(base_size = 10),
  ggplot.component = list(
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
      axis.text.x   = element_text(size = 9)
    )
  )
)
```
:::

**Interpretation:** A significant chi-square result with a higher share of "Bad" bonus scores in new business (NB) than portfolio (P) would confirm that the acquisition channel is less effectively screened than the renewal book. This is a structural pricing problem: new business premiums should incorporate an adverse selection loading relative to renewal premiums for equivalent coverage types.

------------------------------------------------------------------------

### 4.3 Geographic Distribution by Municipality Type and Circulation Area

**Technique justification:** A grouped `geom_col()` chart with policy count annotations is used rather than a pure stacked proportion bar. Both absolute volume and proportional split are analytically important here: the Islands segment's small policy count is as important a fact as its share of the portfolio, because it directly affects the statistical reliability of any downstream loss ratio estimate for that segment. A proportion-only chart would obscure this volume disparity.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-geography
#| fig-cap: "Proportional composition of Urban vs Rural within each municipality type, with absolute policy counts annotated. Stacked bars sum to 100% to allow direct proportion comparison."
#| fig-width: 7
#| fig-height: 5

geo_counts <- insurance_viz |>
  count(municipality_type, circulation_area) |>
  group_by(municipality_type) |>
  mutate(
    pct      = n / sum(n) * 100,
    n_total  = sum(n)
  ) |>
  ungroup()

ggplot(geo_counts,
       aes(x = municipality_type, y = pct, fill = circulation_area)) +
  geom_col(position = "stack", width = 0.6, colour = "white", linewidth = 0.4) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%\n(n=", comma(n), ")")),
    position = position_stack(vjust = 0.5),
    size = 3.2, fontface = "bold", colour = "white"
  ) +
  geom_text(
    data = geo_counts |> distinct(municipality_type, n_total),
    aes(x = municipality_type, y = 103,
        label = paste0("Total:\n", comma(n_total))),
    inherit.aes = FALSE,
    size = 2.8, colour = "grey30", vjust = 0
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    limits = c(0, 115),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("Urban" = "#4B8BBE", "Rural" = "#8E6ABF"),
    name   = "Circulation Area"
  ) +
  labs(
    title    = "Geographic Portfolio Distribution",
    subtitle = "Urban/Rural split within each municipality type; total policy counts above bars",
    x        = "Municipality Type",
    y        = "Proportion (%)"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-geography-code
#| fig-cap: "Proportional composition of Urban vs Rural within each municipality type, with absolute policy counts annotated. Stacked bars sum to 100% to allow direct proportion comparison."
#| fig-width: 7
#| fig-height: 5

geo_counts <- insurance_viz |>
  count(municipality_type, circulation_area) |>
  group_by(municipality_type) |>
  mutate(
    pct      = n / sum(n) * 100,
    n_total  = sum(n)
  ) |>
  ungroup()

ggplot(geo_counts,
       aes(x = municipality_type, y = pct, fill = circulation_area)) +
  geom_col(position = "stack", width = 0.6, colour = "white", linewidth = 0.4) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%\n(n=", comma(n), ")")),
    position = position_stack(vjust = 0.5),
    size = 3.2, fontface = "bold", colour = "white"
  ) +
  geom_text(
    data = geo_counts |> distinct(municipality_type, n_total),
    aes(x = municipality_type, y = 103,
        label = paste0("Total:\n", comma(n_total))),
    inherit.aes = FALSE,
    size = 2.8, colour = "grey30", vjust = 0
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    limits = c(0, 115),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("Urban" = "#4B8BBE", "Rural" = "#8E6ABF"),
    name   = "Circulation Area"
  ) +
  labs(
    title    = "Geographic Portfolio Distribution",
    subtitle = "Urban/Rural split within each municipality type; total policy counts above bars",
    x        = "Municipality Type",
    y        = "Proportion (%)"
  )
```
:::

**Interpretation:** Inland Urban dominates the portfolio by a wide margin. The Islands segment has a markedly smaller count – a fact that becomes analytically critical in Section 10: any observed island loss ratio is based on a thin data sample, and confidence intervals will be correspondingly wide. Pricing decisions for this segment should treat the observed median as an uncertain estimate, not a reliable rate basis.

------------------------------------------------------------------------

### 4.4 Distribution of Vehicle Age

**Technique justification:** `geom_histogram()` with `geom_density()` overlay is the standard choice for a continuous variable examined in isolation. Vehicle age is a moderately right-skewed integer variable; the histogram reveals the empirical shape, binwidth 1 reflects the integer granularity of the data, and the density curve smooths the histogram into a readable trend. A boxplot would suppress distributional shape; a ridgeline plot adds no value without a grouping variable.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-vehicle-age
#| fig-cap: "Distribution of vehicle age. Histogram with density overlay; dashed line marks the portfolio median."
#| fig-width: 8
#| fig-height: 4

med_veh_age <- median(insurance_viz$vehicle_age, na.rm = TRUE)

ggplot(insurance_viz |> filter(!is.na(vehicle_age)),
       aes(x = vehicle_age)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 1, fill = "#8E6ABF",
                 colour = "white", linewidth = 0.2) +
  geom_density(colour = "#E07B39", linewidth = 1.1) +
  geom_vline(xintercept = med_veh_age,
             linetype = "dashed", colour = "grey40") +
  annotate("text",
           x     = med_veh_age + 1,
           y     = 0.025,
           label = paste0("Median: ", round(med_veh_age, 1), " yrs"),
           hjust = 0, size = 3.5, colour = "grey40") +
  labs(
    title    = "Distribution of Vehicle Age",
    subtitle = "Dashed line = portfolio median",
    x        = "Vehicle Age (years)",
    y        = "Density"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-vehicle-age-code
#| fig-cap: "Distribution of vehicle age. Histogram with density overlay; dashed line marks the portfolio median."
#| fig-width: 8
#| fig-height: 4

med_veh_age <- median(insurance_viz$vehicle_age, na.rm = TRUE)

ggplot(insurance_viz |> filter(!is.na(vehicle_age)),
       aes(x = vehicle_age)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 1, fill = "#8E6ABF",
                 colour = "white", linewidth = 0.2) +
  geom_density(colour = "#E07B39", linewidth = 1.1) +
  geom_vline(xintercept = med_veh_age,
             linetype = "dashed", colour = "grey40") +
  annotate("text",
           x     = med_veh_age + 1,
           y     = 0.025,
           label = paste0("Median: ", round(med_veh_age, 1), " yrs"),
           hjust = 0, size = 3.5, colour = "grey40") +
  labs(
    title    = "Distribution of Vehicle Age",
    subtitle = "Dashed line = portfolio median",
    x        = "Vehicle Age (years)",
    y        = "Density"
  )
```
:::

**Interpretation:** The distribution is right-skewed with a mode around 20–25 years. The long tail of very old vehicles (\>40 years) represents a small segment likely insured under basic third-party cover where comprehensive replacement value is not relevant. The modal age band will be revisited in the severity-by-vehicle-age analysis in Section 5.4, where older vehicles show different cost profiles to newer ones.

------------------------------------------------------------------------

### 4.5 Distribution of Vehicle Value

**Technique justification:** The vehicle value distribution is right-skewed and spans two orders of magnitude (€1,630 to €374,000). The log₁₀ scale is applied because insurance loss and asset value data are widely modelled using the lognormal distribution in actuarial practice (Klugman, Panjer & Willmot, 2019; Frees et al., *Loss Data Analytics*; Venegas-Martínez et al., 2014), a distribution whose defining property is that the logarithm of the variable is normally distributed. Plotting on a log scale therefore converts the multiplicative spread of the raw data into an additive spread, revealing the near-normal shape that confirms the lognormal data-generating process (Wilke, 2019).

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-vehicle-value
#| fig-cap: "Distribution of vehicle value on a log₁₀ scale. The near-normal shape confirms a log-normal data-generating process standard for insured asset values."
#| fig-width: 8
#| fig-height: 4

med_veh_val <- median(insurance_viz$vehicle_value, na.rm = TRUE)

ggplot(insurance_viz |> filter(!is.na(vehicle_value)),
       aes(x = vehicle_value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#4B8BBE",
                 colour = "white", linewidth = 0.2) +
  geom_density(colour = "#C0392B", linewidth = 1.1) +
  scale_x_log10(labels = comma) +
  geom_vline(xintercept = med_veh_val,
             linetype = "dashed", colour = "grey40") +
  annotate("text",
           x     = med_veh_val * 1.15,
           y     = 0.25,
           label = paste0("Median: €", comma(round(med_veh_val, 0))),
           hjust = 0, size = 3.5, colour = "grey40") +
  labs(
    title    = "Distribution of Vehicle Value",
    subtitle = "Log₁₀ scale; density overlay confirms log-normal shape",
    x        = "Vehicle Value (€, log scale)",
    y        = "Density"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-vehicle-value-code
#| fig-cap: "Distribution of vehicle value on a log₁₀ scale. The near-normal shape confirms a log-normal data-generating process standard for insured asset values."
#| fig-width: 8
#| fig-height: 4

med_veh_val <- median(insurance_viz$vehicle_value, na.rm = TRUE)

ggplot(insurance_viz |> filter(!is.na(vehicle_value)),
       aes(x = vehicle_value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#4B8BBE",
                 colour = "white", linewidth = 0.2) +
  geom_density(colour = "#C0392B", linewidth = 1.1) +
  scale_x_log10(labels = comma) +
  geom_vline(xintercept = med_veh_val,
             linetype = "dashed", colour = "grey40") +
  annotate("text",
           x     = med_veh_val * 1.15,
           y     = 0.25,
           label = paste0("Median: €", comma(round(med_veh_val, 0))),
           hjust = 0, size = 3.5, colour = "grey40") +
  labs(
    title    = "Distribution of Vehicle Value",
    subtitle = "Log₁₀ scale; density overlay confirms log-normal shape",
    x        = "Vehicle Value (€, log scale)",
    y        = "Density"
  )
```
:::

**Interpretation:** On the log₁₀ scale, the density overlay shows a near-symmetric, bell-shaped curve indicating that vehicle values in this portfolio are spread multiplicatively rather than additively. The log scale is applied because the raw data spans two orders of magnitude (€1,630 to €374,000), a range within which a linear axis would compress the majority of policies into an unreadable sliver at the left edge (Wilke, 2019). The median declared value is approximately €24,700. Notably, vehicle value data is skewed right on the raw scale, which is common for monetary variables where a small number of high-value policies pull the mean substantially above the median, as observed in the data here.

------------------------------------------------------------------------

### 4.6 Distribution of Driver Age

**Technique justification:** `geom_histogram()` + `geom_density()` with vertical reference lines. Driver age is near-symmetric without extreme skewness, so the linear scale is appropriate. Reference lines at ages 25 and 65 mark the boundary age bands used in prior actuarial research on driver risk (Meng et al., 2022; Cappelletti, 2023; ABI, 2009), making the populations below 25 and above 65 immediately visible in the chart without requiring the reader to estimate these boundaries from memory.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-driver-age
#| fig-cap: "Distribution of driver age. Reference lines at 25 and 65 mark the standard actuarial risk segment boundaries."
#| fig-width: 8
#| fig-height: 4

ggplot(insurance_viz, aes(x = driver_age)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 2, fill = "#5C9E6E",
                 colour = "white", linewidth = 0.2) +
  geom_density(colour = "#C0392B", linewidth = 1.1) +
  geom_vline(xintercept = c(25, 65),
             linetype = "dashed", colour = "grey50", linewidth = 0.8) +
  annotate("text", x = 26.5, y = 0.037,
           label = "25\n(young driver\nthreshold)",
           hjust = 0, size = 2.9, colour = "grey50") +
  annotate("text", x = 66.5, y = 0.037,
           label = "65\n(senior\nthreshold)",
           hjust = 0, size = 2.9, colour = "grey50") +
  labs(
    title    = "Distribution of Driver Age",
    subtitle = "Portfolio concentrated in the 35–60 age band",
    x        = "Driver Age (years)",
    y        = "Density"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-driver-age-code
#| fig-cap: "Distribution of driver age. Reference lines at 25 and 65 mark the standard actuarial risk segment boundaries."
#| fig-width: 8
#| fig-height: 4

ggplot(insurance_viz, aes(x = driver_age)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 2, fill = "#5C9E6E",
                 colour = "white", linewidth = 0.2) +
  geom_density(colour = "#C0392B", linewidth = 1.1) +
  geom_vline(xintercept = c(25, 65),
             linetype = "dashed", colour = "grey50", linewidth = 0.8) +
  annotate("text", x = 26.5, y = 0.037,
           label = "25\n(young driver\nthreshold)",
           hjust = 0, size = 2.9, colour = "grey50") +
  annotate("text", x = 66.5, y = 0.037,
           label = "65\n(senior\nthreshold)",
           hjust = 0, size = 2.9, colour = "grey50") +
  labs(
    title    = "Distribution of Driver Age",
    subtitle = "Portfolio concentrated in the 35–60 age band",
    x        = "Driver Age (years)",
    y        = "Density"
  )
```
:::

**Interpretation:** The portfolio is concentrated in the 35–60 age band, with both extremes thinly represented. Young drivers under 25 account for a small share of policies; drivers over 65 taper off gradually. Whether the thin representation of the extremes is due to pricing, voluntary non-participation, or other market factors cannot be determined from this dataset alone. What can be observed directly is that both risk extremes are under-represented relative to the middle-age band; a composition fact that affects the interpretability of age-based analyses in Section 6.1.

------------------------------------------------------------------------

## 5 Bivariate Analysis: Conventional Drivers

### 5.1 Loss Ratio by Policy Type

**Technique justification:** `geom_violin()` with a nested `geom_boxplot()` and Kruskal-Wallis test via `stat_compare_means()`. A violin plot is essential here rather than a plain boxplot: loss ratio is strongly right-skewed and potentially multi-modal within groups. The violin width reveals whether a group's distribution is concentrated (narrow) or dispersed (wide), and whether it is unimodal or carries a secondary mass at higher values – information a boxplot suppresses entirely. The Kruskal-Wallis test is used in preference to ANOVA because loss ratios are non-normally distributed.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-lr-policy-type
#| fig-cap: "Loss ratio distributions by policy type. Violin width encodes density; nested boxplot shows median and IQR. Kruskal-Wallis test confirms groups differ significantly."
#| fig-width: 10
#| fig-height: 5

ggplot(
  insurance_viz |> filter(loss_ratio_capped > 0, total_exposure > 0),
  aes(x = policy_type, y = loss_ratio_capped, fill = policy_type)
) +
  geom_violin(alpha = 0.45, trim = TRUE, show.legend = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA,
               fill = "white", colour = "grey30",
               show.legend = FALSE) +
  stat_compare_means(
    method  = "kruskal.test",
    label.x = 1.2,
    label.y = quantile(insurance_viz$loss_ratio_capped,
                       0.995, na.rm = TRUE) * 0.97
  ) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  annotate("text", x = 0.65, y = 1.04,
           label = "Break-even (LR = 1)",
           hjust = 0, size = 3, colour = "grey50") +
  scale_fill_viridis_d(option = "D") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(
    title    = "Loss Ratio Distribution by Policy Type",
    subtitle = "Violin + boxplot; dashed line = break-even",
    x        = "Policy Type",
    y        = "Loss\nRatio"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-lr-policy-type-code
#| fig-cap: "Loss ratio distributions by policy type. Violin width encodes density; nested boxplot shows median and IQR. Kruskal-Wallis test confirms groups differ significantly."
#| fig-width: 10
#| fig-height: 5

ggplot(
  insurance_viz |> filter(loss_ratio_capped > 0, total_exposure > 0),
  aes(x = policy_type, y = loss_ratio_capped, fill = policy_type)
) +
  geom_violin(alpha = 0.45, trim = TRUE, show.legend = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA,
               fill = "white", colour = "grey30",
               show.legend = FALSE) +
  stat_compare_means(
    method  = "kruskal.test",
    label.x = 1.2,
    label.y = quantile(insurance_viz$loss_ratio_capped,
                       0.995, na.rm = TRUE) * 0.97
  ) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  annotate("text", x = 0.65, y = 1.04,
           label = "Break-even (LR = 1)",
           hjust = 0, size = 3, colour = "grey50") +
  scale_fill_viridis_d(option = "D") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(
    title    = "Loss Ratio Distribution by Policy Type",
    subtitle = "Violin + boxplot; dashed line = break-even",
    x        = "Policy Type",
    y        = "Loss\nRatio"
  )
```
:::

**Interpretation:** The Kruskal-Wallis test confirms statistically significant differences in loss ratio across policy types (p \< 2.2e-16). Third-Party (TP) exhibits the highest median loss ratio of all five types and one of the widest violins — it is both the least profitable on a median basis and highly dispersed. COMP_N (Comprehensive No Excess) shows the most concentrated mass in the 0–2 range with a relatively small IQR and its median sitting just above the break-even line of LR = 1 — meaning the typical COMP_N policy is marginally loss-making, not profitable. Its comparatively narrower violin shape indicates lower volatility than Third-Party or Combined Cover, but the median position above break-even is the more important profitability signal. All five policy types have some policies above the break-even line, confirming no product type is entirely free of loss-making policies.

------------------------------------------------------------------------

### 5.2 Loss Ratio by Bonus Score

**Technique justification:** `stat_halfeye()` from **ggdist** combined with `geom_boxplot()`, with pairwise Wilcoxon tests. The half-eye (asymmetric density + interval) is preferred over a full violin for a three-level ordinal variable: the mirror redundancy of a full violin is unnecessary when comparing three groups, and the half-eye's asymmetric layout gives more horizontal space for readable comparison. Pairwise tests confirm whether the three bonus score groups have statistically different loss ratio distributions; the direction and ordering of those differences are determined by the chart itself, not assumed in advance.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-lr-bonus
#| fig-cap: "Loss ratio by bonus score. Half-eye reveals the full distributional shape; pairwise Wilcoxon tests confirm whether the Good/Neutral/Bad ordering is statistically monotonic."
#| fig-width: 8
#| fig-height: 5

ggplot(
  insurance_viz |> filter(loss_ratio_capped > 0, total_premium > 0),
  aes(x = bonus_score, y = loss_ratio_capped, fill = bonus_score)
) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               fill = "white", colour = "grey30") +
  stat_compare_means(
    comparisons = list(c("Good", "Neutral"),
                       c("Neutral", "Bad"),
                       c("Good", "Bad")),
    method = "wilcox.test",
    label  = "p.signif"
  ) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  scale_fill_manual(
    values = c("Good" = "#5C9E6E",
               "Neutral" = "#F0C27F",
               "Bad" = "#C0392B")
  ) +
  labs(
    title    = "Loss Ratio by Bonus Score",
    subtitle = "Half-eye + boxplot; pairwise Wilcoxon significance annotated",
    x        = "Bonus Score",
    y        = "Loss\nRatio"
  ) +
  theme(legend.position = "none")
```

## Code

```{r}
#| eval: false
#| label: fig-lr-bonus-code
#| fig-cap: "Loss ratio by bonus score. Half-eye reveals the full distributional shape; pairwise Wilcoxon tests confirm whether the Good/Neutral/Bad ordering is statistically monotonic."
#| fig-width: 8
#| fig-height: 5

ggplot(
  insurance_viz |> filter(loss_ratio_capped > 0, total_premium > 0),
  aes(x = bonus_score, y = loss_ratio_capped, fill = bonus_score)
) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               fill = "white", colour = "grey30") +
  stat_compare_means(
    comparisons = list(c("Good", "Neutral"),
                       c("Neutral", "Bad"),
                       c("Good", "Bad")),
    method = "wilcox.test",
    label  = "p.signif"
  ) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  scale_fill_manual(
    values = c("Good" = "#5C9E6E",
               "Neutral" = "#F0C27F",
               "Bad" = "#C0392B")
  ) +
  labs(
    title    = "Loss Ratio by Bonus Score",
    subtitle = "Half-eye + boxplot; pairwise Wilcoxon significance annotated",
    x        = "Bonus Score",
    y        = "Loss\nRatio"
  ) +
  theme(legend.position = "none")
```
:::

**Interpretation:** All three pairwise Wilcoxon tests are statistically significant (Good vs Neutral: p \< 0.0001; Good vs Bad: p \< 0.0001; Neutral vs Bad: p \< 0.01), confirming that the three bonus score groups have meaningfully different loss ratio distributions. However, the direction of the difference is not a simple monotonic increase from Good to Bad. The Good group shows the highest median (approximately 2.0) and the widest spread, with the largest right-tail extension reaching above 12. The Neutral and Bad groups both show lower medians (approximately 1.5) and more compressed distributions. The key distinction between Neutral and Bad appears to be in distribution shape rather than central tendency. The Bad group's boxplot IQR is slightly narrower, but all three groups have policies above the break-even line of LR = 1. This pattern suggests that the Good bonus score group, despite its label, is associated with higher and more variable loss ratios in this dataset; a counterintuitive finding that warrants further investigation rather than being dismissed or over-interpreted from a single chart.

------------------------------------------------------------------------

### 5.3 Claim Severity by Fuel Type

**Technique justification:** `stat_halfeye()` + `geom_boxplot()` on a log₁₀ scale with a Wilcoxon test. Claim severity in this dataset is right-skewed with a long tail of large claim amounts, as visible in the raw distribution. The log₁₀ scale is applied because the observed spread covers several orders of magnitude. Using a linear scale would compress the majority of claims into an unreadable narrow band at the left edge of the chart (Wilke, 2019). Whether the underlying distribution follows a lognormal or another right-skewed family (e.g. gamma, Pareto) is a modelling question beyond the scope of this visual analysis; the log scale is chosen for visual clarity, not to assert a specific parametric form. On the linear scale, the heavy right tail of large claims compresses the bulk of the distribution into an unreadable narrow band. The half-eye on the log scale reveals whether the difference between Diesel and Gasoline is a location shift (systematic median difference) or a tail effect (same median but different extreme values).

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-severity-fuel
#| fig-cap: "Claim severity by fuel type on a log₁₀ scale. Zero-incurred claims (IBNR) are excluded so the axis reflects genuine repair cost distributions. The Wilcoxon test determines whether diesel vehicles incur statistically higher repair costs per incident."
#| fig-width: 7
#| fig-height: 5

ggplot(
  insurance_viz |>
    filter(!is.na(severity_capped),
           !is.na(fuel_type),
           severity_capped > 0),      # exclude zero-incurred / IBNR claims
  aes(x = fuel_type, y = severity_capped, fill = fuel_type)
) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               fill = "white", colour = "grey30") +
  stat_compare_means(
    method      = "wilcox.test",
    label       = "p.format",
    label.x     = 1.5,
    label.y.npc = 0.9
  ) +
  scale_y_log10(
    labels = comma,
    breaks = c(10, 50, 100, 500, 1000, 5000),
    limits = c(10, 7000)
  ) +
  scale_fill_manual(
    values = c("Diesel" = "#4B8BBE", "Gasoline" = "#E07B39")
  ) +
  labs(
    title    = "Claim Severity by Fuel Type",
    subtitle = "Log₁₀ scale; zero-incurred claims excluded; Wilcoxon test for group difference",
    x        = "Fuel Type",
    y        = "Claim Severity\n(€, log scale)"
  ) +
  theme(legend.position = "none")
```

## Code

```{r}
#| eval: false
#| label: fig-severity-fuel-code
#| fig-cap: "Claim severity by fuel type on a log₁₀ scale. Zero-incurred claims (IBNR) are excluded so the axis reflects genuine repair cost distributions. The Wilcoxon test determines whether diesel vehicles incur statistically higher repair costs per incident."
#| fig-width: 7
#| fig-height: 5

ggplot(
  insurance_viz |>
    filter(!is.na(severity_capped),
           !is.na(fuel_type),
           severity_capped > 0),      # exclude zero-incurred / IBNR claims
  aes(x = fuel_type, y = severity_capped, fill = fuel_type)
) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               fill = "white", colour = "grey30") +
  stat_compare_means(
    method      = "wilcox.test",
    label       = "p.format",
    label.x     = 1.5,
    label.y.npc = 0.9
  ) +
  scale_y_log10(
    labels = comma,
    breaks = c(10, 50, 100, 500, 1000, 5000),
    limits = c(10, 7000)
  ) +
  scale_fill_manual(
    values = c("Diesel" = "#4B8BBE", "Gasoline" = "#E07B39")
  ) +
  labs(
    title    = "Claim Severity by Fuel Type",
    subtitle = "Log₁₀ scale; zero-incurred claims excluded; Wilcoxon test for group difference",
    x        = "Fuel Type",
    y        = "Claim Severity\n(€, log scale)"
  ) +
  theme(legend.position = "none")
```
:::

**Interpretation:** The Wilcoxon test confirms a statistically significant difference in claim severity between diesel and gasoline vehicles (p = 0.04). However, looking at the median lines within the boxplots, Gasoline shows a slightly higher median severity than Diesel on the log scale — the median line for Gasoline sits marginally above that of Diesel. Both distributions span a similar range from approximately €10 to above €5,000, and both show the bulk of density concentrated between €300 and slightly above €1,000 with a secondary cluster near €100. The Gasoline half-eye appears to extend slightly further rightward in the upper density region above €1,000. The Wilcoxon test confirms the two distributions differ statistically, but the direction and practical magnitude of that difference as well as its mechanism, cannot be determined from this chart alone. The vehicle type composition of each fuel group (e.g. whether diesel vehicles in this portfolio are systematically larger or higher-value) is not contained in this dataset and would require additional variables to investigate.

------------------------------------------------------------------------

### 5.4 Claim Severity by Vehicle Age

**Technique justification:** `geom_hex()` + `geom_smooth(method = "gam")` with a log₁₀ y-axis. With 354,140 records, a standard scatter plot produces an uninterpretable mass of overplotted points. The hexbin resolves overplotting by binning points into hexagonal cells coloured by count density, revealing where observations concentrate. The GAM smoother over this density reveals the non-linear age–severity relationship – a relationship that a linear smoother or binned bar chart would distort by imposing functional form assumptions.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-severity-vehicle-age
#| fig-cap: "Claim severity vs vehicle age. Hexbin reveals data density; GAM smoother with 95% CI surfaces the non-linear severity–age relationship."
#| fig-width: 9
#| fig-height: 5

ggplot(
  insurance_viz |>
    filter(!is.na(vehicle_age), !is.na(severity_capped),
           severity_capped > 0),    # exclude zero-incurred claims
  aes(x = vehicle_age, y = severity_capped)
) +
  geom_hex(bins = 50, aes(fill = after_stat(log(count)))) +
  geom_smooth(method  = "gam",
              formula = y ~ s(x, bs = "cs"),
              colour  = "#C0392B", linewidth = 1.2, se = TRUE) +
  scale_fill_viridis_c(name = "log(count)", option = "magma") +
  scale_y_log10(
    labels = comma,
    breaks = c(10, 50, 100, 500, 1000, 5000),
    limits = c(10, 7500)
  ) +
  labs(
    title    = "Claim Severity vs Vehicle Age",
    subtitle = "Hexbin density + GAM smooth with 95% CI; y-axis log scaled",
    x        = "Vehicle Age (years)",
    y        = "Claim Severity\n(€, log scale)"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-severity-vehicle-age-code
#| fig-cap: "Claim severity vs vehicle age. Hexbin reveals data density; GAM smoother with 95% CI surfaces the non-linear severity–age relationship."
#| fig-width: 9
#| fig-height: 5

ggplot(
  insurance_viz |>
    filter(!is.na(vehicle_age), !is.na(severity_capped),
           severity_capped > 0),    # exclude zero-incurred claims
  aes(x = vehicle_age, y = severity_capped)
) +
  geom_hex(bins = 50, aes(fill = after_stat(log(count)))) +
  geom_smooth(method  = "gam",
              formula = y ~ s(x, bs = "cs"),
              colour  = "#C0392B", linewidth = 1.2, se = TRUE) +
  scale_fill_viridis_c(name = "log(count)", option = "magma") +
  scale_y_log10(
    labels = comma,
    breaks = c(10, 50, 100, 500, 1000, 5000),
    limits = c(10, 7500)
  ) +
  labs(
    title    = "Claim Severity vs Vehicle Age",
    subtitle = "Hexbin density + GAM smooth with 95% CI; y-axis log scaled",
    x        = "Vehicle Age (years)",
    y        = "Claim Severity\n(€, log scale)"
  )
```
:::

## **Interpretation:** The hexbin chart confirms that claim severity observations are distributed across vehicle ages from approximately 0 to 60 years, with the highest data density concentrated between ages 5 and 50 (brightest yellow-white hexbin region). The GAM smoother runs largely horizontally across the chart, indicating that claim severity does not change strongly or systematically with vehicle age across the bulk of the distribution. A marginal upward trend is visible at the far right edge of the chart (vehicles aged 50+), consistent with elevated repair costs for very old vehicles where specialist parts may be scarce — but the confidence interval at this extreme is wide, reflecting small policy volumes. The claimed U-shaped pattern (high severity for new vehicles, declining through middle ages, uptick for old vehicles) is not clearly visible in the GAM line as rendered; the relationship is closer to flat with a marginal late-age uptick.

## 6 Bivariate Analysis: Demographic Drivers

### 6.1 Loss Ratio Distributions Across Driver Age Bands

**Technique justification:** `stat_density_ridges()` from **ggridges** with quantile shading (`quantiles = 4`, `calc_ecdf = TRUE`). Grouped boxplots for six age bands would be compact but would hide within-group distributional shape. Prior research on driver age and claim rates (Meng et al., 2022; Cappelletti, 2023; ABI, 2009) suggests elevated loss ratios at young and old age extremes. Whether this pattern is visible in this dataset, and to what degree, is the question the ridgeline chart investigates. Ridgelines with quartile shading communicate this directly and enable visual comparison of distributional shape, not just central tendency.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-lr-age-bands
#| fig-cap: "Loss ratio distributions across driver age bands with quartile shading. The ridgeline format reveals distributional shape differences that a grouped boxplot would suppress."
#| fig-width: 9
#| fig-height: 6

ggplot(
  insurance_viz |>
    filter(loss_ratio_capped > 0, total_exposure > 0,
           !is.na(driver_age_band)),
  aes(x    = loss_ratio_capped,
      y    = driver_age_band,
      fill = factor(after_stat(quantile)))
) +
  stat_density_ridges(
    geom           = "density_ridges_gradient",
    calc_ecdf      = TRUE,
    quantiles      = 4,
    quantile_lines = TRUE,
    scale          = 2.2,
    rel_min_height = 0.005,
    bandwidth      = 0.04
  ) +
  scale_fill_viridis_d(name = "Quartile", option = "D", alpha = 0.85) +
  scale_x_continuous(
    limits = c(0, 6),
    labels = number_format(accuracy = 0.1),
    expand = c(0, 0)
  ) +
  scale_y_discrete(expand = expansion(add = c(0.3, 2.5))) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  labs(
    title    = "Loss Ratio Distribution by Driver Age Band",
    subtitle = "Ridgeline with quartile shading; dashed line = break-even (LR = 1)",
    x        = "Loss Ratio (capped at 99th percentile)",
    y        = "Driver\nAge Band"
  ) +
  theme_ridges(font_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(hjust = 0.5, colour = "grey40"),
    legend.position = "right"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-lr-age-bands-code
#| fig-cap: "Loss ratio distributions across driver age bands with quartile shading. The ridgeline format reveals distributional shape differences that a grouped boxplot would suppress."
#| fig-width: 9
#| fig-height: 6

ggplot(
  insurance_viz |>
    filter(loss_ratio_capped > 0, total_exposure > 0,
           !is.na(driver_age_band)),
  aes(x    = loss_ratio_capped,
      y    = driver_age_band,
      fill = factor(after_stat(quantile)))
) +
  stat_density_ridges(
    geom           = "density_ridges_gradient",
    calc_ecdf      = TRUE,
    quantiles      = 4,
    quantile_lines = TRUE,
    scale          = 2.2,
    rel_min_height = 0.005,
    bandwidth      = 0.04
  ) +
  scale_fill_viridis_d(name = "Quartile", option = "D", alpha = 0.85) +
  scale_x_continuous(
    limits = c(0, 6),
    labels = number_format(accuracy = 0.1),
    expand = c(0, 0)
  ) +
  scale_y_discrete(expand = expansion(add = c(0.3, 2.5))) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  labs(
    title    = "Loss Ratio Distribution by Driver Age Band",
    subtitle = "Ridgeline with quartile shading; dashed line = break-even (LR = 1)",
    x        = "Loss Ratio (capped at 99th percentile)",
    y        = "Driver\nAge Band"
  ) +
  theme_ridges(font_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(hjust = 0.5, colour = "grey40"),
    legend.position = "right"
  )
```
:::

**Interpretation:** The 18–25 age band shows the widest distribution and the heaviest right skew among all six bands with a larger share of policies in this group producing high loss ratios compared to middle-age bands. The 36–55 bands show the most compact distributions with the lowest median loss ratios. The 65+ band shows a moderately wider distribution than the 36–55 group and a slightly elevated median. These observed differences in distributional shape and spread are consistent with the patterns reported in prior driver risk research (Meng et al., 2022; Cappelletti, 2023), though caution is warranted given the thin representation of the under-25 and over-65 segments in this portfolio. The estimates for those bands carry more uncertainty than those for the well-populated middle-age bands.

------------------------------------------------------------------------

### 6.2 Older Driver + Young Car Interaction

**Technique justification:** `geom_tile()` heatmap with `geom_text()` cell labels. The analytical question requires examining the *interaction* of two binned variables against median claim frequency. A heatmap is the most direct representation of a two-dimensional cell matrix: colour fills communicate the KPI simultaneously across all cells, while text annotations show the policy count – enabling the reader to distinguish data-rich cells (trustworthy estimates) from data-sparse cells (high uncertainty). A faceted bar chart or grouped boxplot would be unreadable with six × six = thirty-six combinations.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-older-driver-young-car
#| fig-cap: "Mean claim frequency heatmap: driver age × vehicle age interaction. Red cells = higher claim rates. Mean is used instead of median because the portfolio is zero-inflated (median is 0 for every cell). Cells with n < 30 are excluded as statistically unreliable."
#| fig-width: 10
#| fig-height: 6

heatmap_data <- insurance_viz |>
  filter(!is.na(driver_age_band), !is.na(vehicle_age_band),
         total_exposure > 0) |>
  group_by(driver_age_band, vehicle_age_band) |>
  summarise(
    mean_freq = mean(claim_frequency, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  ) |>
  filter(n >= 30) |>                          # suppress unreliable low-n cells
  mutate(mean_freq = pmin(mean_freq, 2.0))    # cap one anomalous outlier cell (n=3)

mid_val <- median(heatmap_data$mean_freq, na.rm = TRUE)

ggplot(heatmap_data,
       aes(x = vehicle_age_band,
           y = driver_age_band,
           fill = mean_freq)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(
    aes(label = paste0(round(mean_freq, 2),
                       "\n(n=", comma(n), ")"),
        colour = mean_freq > mid_val),        # white text on dark cells, dark on light
    size = 2.6, fontface = "bold", show.legend = FALSE
  ) +
  scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "grey20")) +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "#F7F7F7",
    high     = "#D6604D",
    midpoint = mid_val,
    name     = "Mean\nClaim Freq",
    labels   = number_format(accuracy = 0.01)
  ) +
  labs(
    title    = "Claim Frequency: Driver Age × Vehicle Age Interaction",
    subtitle = "Mean claims per policy year; darker red = higher claim rate; cells with n < 30 excluded",
    x        = "Vehicle Age Band (years)",
    y        = "Driver\nAge Band"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Code

```{r}
#| eval: false
#| label: fig-older-driver-young-car-code
#| fig-cap: "Mean claim frequency heatmap: driver age × vehicle age interaction. Red cells = higher claim rates. Mean is used instead of median because the portfolio is zero-inflated (median is 0 for every cell). Cells with n < 30 are excluded as statistically unreliable."
#| fig-width: 10
#| fig-height: 6

heatmap_data <- insurance_viz |>
  filter(!is.na(driver_age_band), !is.na(vehicle_age_band),
         total_exposure > 0) |>
  group_by(driver_age_band, vehicle_age_band) |>
  summarise(
    mean_freq = mean(claim_frequency, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  ) |>
  filter(n >= 30) |>                          # suppress unreliable low-n cells
  mutate(mean_freq = pmin(mean_freq, 2.0))    # cap one anomalous outlier cell (n=3)

mid_val <- median(heatmap_data$mean_freq, na.rm = TRUE)

ggplot(heatmap_data,
       aes(x = vehicle_age_band,
           y = driver_age_band,
           fill = mean_freq)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(
    aes(label = paste0(round(mean_freq, 2),
                       "\n(n=", comma(n), ")"),
        colour = mean_freq > mid_val),        # white text on dark cells, dark on light
    size = 2.6, fontface = "bold", show.legend = FALSE
  ) +
  scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "grey20")) +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "#F7F7F7",
    high     = "#D6604D",
    midpoint = mid_val,
    name     = "Mean\nClaim Freq",
    labels   = number_format(accuracy = 0.01)
  ) +
  labs(
    title    = "Claim Frequency: Driver Age × Vehicle Age Interaction",
    subtitle = "Mean claims per policy year; darker red = higher claim rate; cells with n < 30 excluded",
    x        = "Vehicle Age Band (years)",
    y        = "Driver\nAge Band"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
:::

**Interpretation:** The heatmap reveals two notable patterns. First, the 18–25 × 4–7 cell stands out as the darkest cell in the entire chart with a mean claim frequency of 0.80 (n=1,296) — substantially higher than all adjacent cells and all other age bands at the same vehicle age. This is the single most prominent signal in the heatmap. Second, across the middle driver age bands (26–55), claim frequency is consistently higher for newer vehicles (0–3 and 4–7 year bands, showing values of 0.52–0.58) and declines as vehicle age increases, converging toward approximately 0.29–0.32 for vehicles aged 25+ years. The 65+ row does not show a clear pattern of elevated frequency for newer vehicles: values range from 0.20 to 0.32 across vehicle age bands without a consistent directional trend, and the small cell sizes (n=70 to n=176 for newer vehicle bands) mean these estimates carry substantial uncertainty and should not be over-interpreted. Cells with n \< 200 are particularly unreliable as standalone estimates.

------------------------------------------------------------------------

## 7 Bivariate Analysis: Brand and Status Drivers

### 7.1 Do Prestige Brand Drivers Make Fewer Claims – or Just Different Ones?

**Technique justification:** Side-by-side `stat_pointinterval()` panels from **ggdist** – one for claim frequency, one for claim severity – composed with **patchwork**. A single loss ratio chart conflates two fundamentally different risk dimensions: frequency (how often) and severity (how much each time). The hypothesis here is specifically that luxury brands may show *low frequency but high severity* – a divergence that is invisible in a combined loss ratio. Presenting both dimensions side-by-side with median + CI makes this decomposition explicit.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-brand-freq-severity
#| fig-cap: "Claim frequency (left) and claim severity (right) by brand tier. Side-by-side panels decompose two distinct risk dimensions that a single loss ratio chart conflates."
#| fig-width: 11
#| fig-height: 5

brand_data <- insurance_viz |>
  filter(!is.na(brand_tier), total_exposure > 0)

p_freq <- ggplot(brand_data,
                 aes(x = brand_tier, y = claim_frequency)) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 3,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    name   = "CI",
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95%", "99%")
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(title = "Claim frequency by brand tier",
       x = "Brand Tier", y = "Claim\nFrequency")

p_sev <- ggplot(
  brand_data |> filter(!is.na(severity_capped), severity_capped > 0),
  aes(x = brand_tier, y = severity_capped)
) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 3,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    name   = "CI",
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95%", "99%")
  ) +
  scale_y_log10(
    labels = comma,
    breaks = c(10, 50, 100, 500, 1000, 5000),
    limits = c(10, 7500)
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(title = "Claim severity by brand tier",
       x = "Brand Tier", y = "Claim Severity\n(€, log scale)")

p_freq + p_sev +
  plot_annotation(
    title    = "Brand Tier: Frequency vs Severity Decomposition",
    subtitle = "Median point + 95%/99% confidence intervals",
    theme    = theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40")
    )
  ) &
  theme(legend.position = "right")
```

## Code

```{r}
#| eval: false
#| label: fig-brand-freq-severity-code
#| fig-cap: "Claim frequency (left) and claim severity (right) by brand tier. Side-by-side panels decompose two distinct risk dimensions that a single loss ratio chart conflates."
#| fig-width: 11
#| fig-height: 5

brand_data <- insurance_viz |>
  filter(!is.na(brand_tier), total_exposure > 0)

p_freq <- ggplot(brand_data,
                 aes(x = brand_tier, y = claim_frequency)) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 3,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    name   = "CI",
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95%", "99%")
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(title = "Claim frequency by brand tier",
       x = "Brand Tier", y = "Claim\nFrequency")

p_sev <- ggplot(
  brand_data |> filter(!is.na(severity_capped), severity_capped > 0),
  aes(x = brand_tier, y = severity_capped)
) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 3,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    name   = "CI",
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95%", "99%")
  ) +
  scale_y_log10(
    labels = comma,
    breaks = c(10, 50, 100, 500, 1000, 5000),
    limits = c(10, 7500)
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(title = "Claim severity by brand tier",
       x = "Brand Tier", y = "Claim Severity\n(€, log scale)")

p_freq + p_sev +
  plot_annotation(
    title    = "Brand Tier: Frequency vs Severity Decomposition",
    subtitle = "Median point + 95%/99% confidence intervals",
    theme    = theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40")
    )
  ) &
  theme(legend.position = "right")
```
:::

**Interpretation:** The two panels together reveal a clear and commercially significant pattern. On claim frequency (left), all three brand tiers show near-identical median values close to zero with overlapping confidence intervals – meaning brand tier does **not** discriminate how often claims occur. The "prestige paradox" hypothesis of luxury drivers claiming less frequently is **not supported**: frequency is statistically indistinguishable across tiers. However, the severity panel (right) tells the opposite story. All three tiers show median severity around €500–650, but the confidence intervals diverge substantially at the upper end. Luxury brands show a notably tighter 95% CI band in the lower range, while the 99% CI extends further upward than the other tiers – confirming that while the *typical* luxury claim costs a similar amount to a mass-market claim, the *extreme* claims from luxury vehicles are disproportionately costly. For an underwriter, the actionable implication is: brand tier loading should be applied **not** to frequency (which is uniform) but as a severity tail loading, particularly for catastrophic claims. The risk of under-reserving luxury vehicle large losses is real and not captured by a combined loss ratio view.

------------------------------------------------------------------------

### 7.2 The "Bad Bonus, Fancy Car" Paradox

**Technique justification:** `ggbarstats()` with brand tier on the y-axis and bonus score as the fill variable. The chi-square test directly tests whether bonus score distribution is independent of brand tier – this is the key statistical question. A plain grouped bar chart would require the reader to visually judge differences and infer significance; `ggbarstats()` embeds the test in the chart and provides an exact p-value, making the analytical conclusion unambiguous.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-bad-bonus-brand
#| fig-cap: "Bonus score composition by brand tier. Hover each segment for counts and percentages. χ² p < 0.001 (χ²=281.9, df=4) confirms significant association."

brand_bonus_data <- insurance_viz |>
  filter(!is.na(brand_tier)) |>
  count(brand_tier, bonus_score) |>
  group_by(brand_tier) |>
  mutate(
    pct     = n / sum(n) * 100,
    n_total = sum(n),
    tooltip = paste0(
      "<b>", brand_tier, "</b><br>",
      "Bonus: <b>", bonus_score, "</b><br>",
      "Count: ", scales::comma(n), "<br>",
      "Share: ", round(pct, 1), "%<br>",
      "Tier total: ", scales::comma(n_total)
    )
  ) |>
  ungroup() |>
  mutate(
    bonus_score = factor(bonus_score, levels = c("Good","Neutral","Bad")),
    brand_tier  = factor(brand_tier,
      levels = c("Luxury","Upper-mid / Mass-market","Budget / Commercial"))
  )

p_brand <- ggplot(brand_bonus_data,
                  aes(x = brand_tier, y = pct, fill = bonus_score,
                      text = tooltip)) +
  geom_col(position = "stack", width = 0.55, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = bonus_colours, name = "Bonus Score") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  labs(
    title    = "Bonus Score Composition by Vehicle Brand Tier",
    subtitle = "χ²=281.9 (df=4), p < 0.001 – Luxury tier carries the highest Bad bonus proportion",
    x = "Brand Tier", y = "Proportion (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, colour = "grey40", size = 9))

ggplotly(p_brand, tooltip = "text") |>
  layout(legend = list(title = list(text = "<b>Bonus Score</b>")))
```

## Code

```{r}
#| eval: false
#| label: fig-bad-bonus-brand-code
#| fig-cap: "Bonus score composition by brand tier. The Pearson χ² test confirms whether luxury brand policyholders carry a systematically different claims history profile."
#| fig-width: 8
#| fig-height: 6

ggbarstats(
  data             = insurance_viz |> filter(!is.na(brand_tier)),
  x                = bonus_score,
  y                = brand_tier,
  label            = "both",
  xlab             = "Brand Tier",
  ylab             = "Proportion",
  legend.title     = "Bonus Score",
  title            = "Bonus Score Composition by Vehicle Brand Tier",
  subtitle         = "Pearson χ²: does risk profile differ significantly by brand tier?",
  palette          = "Set2",
  ggtheme          = theme_minimal(base_size = 10),
  ggplot.component = list(
    scale_fill_manual(
      values = c("Good" = "#5C9E6E", "Neutral" = "#F0C27F", "Bad" = "#C0392B"),
      name   = "Bonus Score"
    ),
    scale_x_discrete(labels = function(x) str_wrap(x, width = 14)),
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
      axis.text.x   = element_text(size = 9)
    )
  )
)
```
:::

**Interpretation:** The chi-square test is highly significant (χ² = 281.9, df = 4, p \< 0.001), confirming that bonus score distribution differs meaningfully across brand tiers. Examining the actual proportions: Luxury tier carries the highest "Bad" bonus proportion at 2.0%, compared to 1.3% for Budget/Commercial and 1.2% for Upper-mid/Mass-market. While the absolute differences are modest, they are statistically robust given the large sample sizes. More notable is the "Neutral" score proportion: Luxury holds 2.7% Neutral policyholders versus 2.1% for Upper-mid/Mass-market – a 29% relative difference. This means Luxury vehicles are held by a somewhat higher proportion of policyholders with Neutral or Bad bonus scores than other brand tiers, a statistically confirmed difference, though modest in absolute magnitude. The chi-square effect size (Cramér's V) is small, indicating the practical difference between tiers is limited. Whether this observed association warrants a separate brand tier loading in a pricing model is a business decision that would require further multivariate analysis. This dataset can identify the association but cannot determine whether brand tier carries independent predictive power once other variables (vehicle value, bonus score) are controlled for.

------------------------------------------------------------------------

### 7.3 Power-to-Weight Ratio as a Latent Thrill-Seeker Index

**Technique justification:** `geom_hex()` + `geom_smooth(method = "gam")` faceted by brand tier. The faceting tests a specific sub-hypothesis: is the power-to-weight signal stronger within the Upper-mid / Mass-market tier (where accessible performance cars concentrate) than within the Luxury tier (where even high-performance vehicles are driven cautiously)? If the GAM slope is steeper in the mass-market panel, this confirms the ratio as a valid additional rating factor for non-luxury segments specifically.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-power-weight
#| fig-cap: "Claim frequency vs power-to-weight ratio by brand tier. Faceted GAM curves test whether the performance-proxy signal differs across tiers."
#| fig-width: 11
#| fig-height: 5

pw_cap <- quantile(insurance_viz$power_to_weight_ratio, 0.99, na.rm = TRUE)

ggplot(
  insurance_viz |>
    filter(!is.na(brand_tier), total_exposure > 0,
           claim_frequency < 10,
           power_to_weight_ratio < pw_cap),
  aes(x = power_to_weight_ratio, y = claim_frequency)
) +
  geom_hex(bins = 40, aes(fill = after_stat(log(count)))) +
  geom_smooth(method  = "gam",
              formula = y ~ s(x, bs = "cs"),
              colour  = "#C0392B", linewidth = 1.1, se = TRUE) +
  scale_fill_viridis_c(option = "magma", name = "log(count)") +
  facet_wrap(~brand_tier, nrow = 1) +
  labs(
    title    = "Claim Frequency vs Power-to-Weight Ratio by Brand Tier",
    subtitle = "GAM smoother with 95% CI; x-axis capped at 99th percentile",
    x        = "Power-to-Weight Ratio (kg/hp)",
    y        = "Claim\nFrequency"
  ) +
  theme(strip.text = element_text(size = 10, face = "bold"))
```

## Code

```{r}
#| eval: false
#| label: fig-power-weight-code
#| fig-cap: "Claim frequency vs power-to-weight ratio by brand tier. Faceted GAM curves test whether the performance-proxy signal differs across tiers."
#| fig-width: 11
#| fig-height: 5

pw_cap <- quantile(insurance_viz$power_to_weight_ratio, 0.99, na.rm = TRUE)

ggplot(
  insurance_viz |>
    filter(!is.na(brand_tier), total_exposure > 0,
           claim_frequency < 10,
           power_to_weight_ratio < pw_cap),
  aes(x = power_to_weight_ratio, y = claim_frequency)
) +
  geom_hex(bins = 40, aes(fill = after_stat(log(count)))) +
  geom_smooth(method  = "gam",
              formula = y ~ s(x, bs = "cs"),
              colour  = "#C0392B", linewidth = 1.1, se = TRUE) +
  scale_fill_viridis_c(option = "magma", name = "log(count)") +
  facet_wrap(~brand_tier, nrow = 1) +
  labs(
    title    = "Claim Frequency vs Power-to-Weight Ratio by Brand Tier",
    subtitle = "GAM smoother with 95% CI; x-axis capped at 99th percentile",
    x        = "Power-to-Weight Ratio (kg/hp)",
    y        = "Claim\nFrequency"
  ) +
  theme(strip.text = element_text(size = 10, face = "bold"))
```
:::

**Interpretation:** The three faceted panels reveal that the GAM smoother is effectively flat across all brand tiers – there is no meaningful monotonically increasing relationship between power-to-weight ratio and claim frequency in any tier. The hexbin density confirms that most policies concentrate between power-to-weight ratios of 5–15 kg/hp, with claim frequencies predominantly at zero. The wide confidence bands across all x-values indicate high uncertainty in the smoother, driven by the sparse data at extreme power-to-weight values. **The "thrill-seeker" hypothesis is not supported:** power-to-weight ratio does not discriminate claim frequency after accounting for brand tier in this portfolio. Two explanations are plausible: first, the portfolio is dominated by mainstream vehicles where power-to-weight variation is narrow and claims are driven by other factors; second, genuine high-performance vehicles (low power-to-weight ratios indicating very high power per kg) are too few in this dataset to generate a detectable signal. **For the underwriter, this means power-to-weight ratio should not be introduced as a standalone rating factor** – it would add model complexity without improving risk differentiation. The variable may have value in a multivariate model, but as a univariate predictor it is not credible in this portfolio.

------------------------------------------------------------------------

### 7.4 Which Brands Are Most Expensive to Repair Relative to Declared Value?

**Technique justification:** Ranked `stat_pointinterval()` dotplot, restricted to brands with at least 200 claimant policies. A repair cost intensity index (claim severity / vehicle value) captures the mismatch between what was declared for pricing and what repairs actually cost. The ranked dotplot with CI bars is used in preference to a bar chart because it simultaneously communicates the point estimate and the estimation uncertainty: a bar chart of medians would convey false precision for low-volume brands.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-repair-intensity
#| fig-cap: "Repair cost intensity (claim severity / vehicle value) by vehicle brand. Brands with high intensity relative to their declared value are the portfolio's pricing blind spots."
#| fig-width: 10
#| fig-height: 7

repair_data <- insurance_viz |>
  filter(total_claims > 0, !is.na(vehicle_value),
         vehicle_value > 0, !is.na(severity_capped)) |>
  mutate(repair_intensity = severity_capped / vehicle_value) |>
  group_by(vehicle_brand) |>
  filter(n() >= 200) |>
  ungroup()

brand_order <- repair_data |>
  group_by(vehicle_brand) |>
  summarise(med = median(repair_intensity, na.rm = TRUE)) |>
  arrange(med) |>
  pull(vehicle_brand)

repair_data |>
  mutate(vehicle_brand = factor(vehicle_brand, levels = brand_order)) |>
  ggplot(aes(x = vehicle_brand, y = repair_intensity)) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 2.5,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95% CI", "99% CI"),
    name   = "CI"
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed",
             colour = "grey50", linewidth = 0.6) +
  annotate("text", x = 1, y = 0.52,
           label = "50% of declared value",
           hjust = 0, size = 3, colour = "grey50") +
  coord_flip() +
  labs(
    title    = "Repair Cost Intensity by Vehicle Brand",
    subtitle = "Index = claim severity / vehicle value; brands with ≥200 claimant policies",
    x        = "Vehicle Brand",
    y        = "Repair Cost Intensity\n(Severity / Declared Value)"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-repair-intensity-code
#| fig-cap: "Repair cost intensity (claim severity / vehicle value) by vehicle brand. Brands with high intensity relative to their declared value are the portfolio's pricing blind spots."
#| fig-width: 10
#| fig-height: 7

repair_data <- insurance_viz |>
  filter(total_claims > 0, !is.na(vehicle_value),
         vehicle_value > 0, !is.na(severity_capped)) |>
  mutate(repair_intensity = severity_capped / vehicle_value) |>
  group_by(vehicle_brand) |>
  filter(n() >= 200) |>
  ungroup()

brand_order <- repair_data |>
  group_by(vehicle_brand) |>
  summarise(med = median(repair_intensity, na.rm = TRUE)) |>
  arrange(med) |>
  pull(vehicle_brand)

repair_data |>
  mutate(vehicle_brand = factor(vehicle_brand, levels = brand_order)) |>
  ggplot(aes(x = vehicle_brand, y = repair_intensity)) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 2.5,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95% CI", "99% CI"),
    name   = "CI"
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed",
             colour = "grey50", linewidth = 0.6) +
  annotate("text", x = 1, y = 0.52,
           label = "50% of declared value",
           hjust = 0, size = 3, colour = "grey50") +
  coord_flip() +
  labs(
    title    = "Repair Cost Intensity by Vehicle Brand",
    subtitle = "Index = claim severity / vehicle value; brands with ≥200 claimant policies",
    x        = "Vehicle Brand",
    y        = "Repair Cost Intensity\n(Severity / Declared Value)"
  )
```
:::

**Interpretation:** Brands with high repair cost intensity and narrow confidence intervals (data-rich) represent confirmed pricing blind spots: their premiums are calibrated to low declared values but their actual repair costs are disproportionately high. Older luxury brands that have depreciated significantly in declared value while retaining expensive parts and specialist labour requirements are the typical culprits. Brands whose entire 99% CI sits above 0.5 are systematically generating repair costs that consume more than half of the declared vehicle value per incident – a clear signal for a brand-specific severity surcharge.

------------------------------------------------------------------------

## 8 Bivariate Analysis: Behavioural Drivers

### 8.1 The Serial Claimer – Does Multi-Line Claiming Predict Future Total Claims?

**Technique justification:** `geom_tile()` heatmap of claim breadth (distinct claim types in a year) × bonus score, with policy count annotations. The heatmap is preferred over a cross-tabulation table or grouped bar chart because it exposes the *joint* distribution across both dimensions simultaneously, and anomalous cells – particularly "claim breadth ≥ 3 with Good bonus score" – visually stand out as unexpectedly warm in an otherwise cool region of the matrix.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-serial-claimer
#| fig-cap: "Heatmap: claim breadth (distinct claim types) vs bonus score for claimant policies. Anomalous hot cells in the high-breadth / Good-bonus region flag potential mis-scoring."
#| fig-width: 8
#| fig-height: 5

serial_data <- insurance_viz |>
  filter(total_claims > 0) |>
  count(claim_breadth, bonus_score) |>
  group_by(bonus_score) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(claim_breadth = factor(claim_breadth))

ggplot(serial_data,
       aes(x = bonus_score, y = claim_breadth, fill = pct)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%\n(n=", comma(n), ")")),
    size   = 3,
    colour = ifelse(serial_data$pct > 25, "white", "grey20"),
    fontface = "bold"
  ) +
  scale_fill_gradient2(
    low      = "#E6F1FB",
    mid      = "#FAC775",
    high     = "#C0392B",
    midpoint = 20,
    name     = "% within\nbonus group"
  ) +
  labs(
    title    = "Claim Breadth by Bonus Score (Claimant Policies Only)",
    subtitle = "High breadth + Good bonus = anomalous cells worth investigating",
    x        = "Bonus Score",
    y        = "Claim Breadth\n(distinct claim types in year)"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-serial-claimer-code
#| fig-cap: "Heatmap: claim breadth (distinct claim types) vs bonus score for claimant policies. Anomalous hot cells in the high-breadth / Good-bonus region flag potential mis-scoring."
#| fig-width: 8
#| fig-height: 5

serial_data <- insurance_viz |>
  filter(total_claims > 0) |>
  count(claim_breadth, bonus_score) |>
  group_by(bonus_score) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(claim_breadth = factor(claim_breadth))

ggplot(serial_data,
       aes(x = bonus_score, y = claim_breadth, fill = pct)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%\n(n=", comma(n), ")")),
    size   = 3,
    colour = ifelse(serial_data$pct > 25, "white", "grey20"),
    fontface = "bold"
  ) +
  scale_fill_gradient2(
    low      = "#E6F1FB",
    mid      = "#FAC775",
    high     = "#C0392B",
    midpoint = 20,
    name     = "% within\nbonus group"
  ) +
  labs(
    title    = "Claim Breadth by Bonus Score (Claimant Policies Only)",
    subtitle = "High breadth + Good bonus = anomalous cells worth investigating",
    x        = "Bonus Score",
    y        = "Claim Breadth\n(distinct claim types in year)"
  )
```
:::

**Interpretation:** Most claimants file under a single coverage line, which is expected. The key anomalous finding would be Good-bonus policyholders accounting for a disproportionate share of breadth-3+ claimants: their complex multi-line claims are inconsistent with a "Good" risk classification. These policyholders either experienced genuinely independent adverse events in one year (unlikely for 3+ distinct lines simultaneously), or represent a segment where the bonus scoring methodology is lagged and fails to capture complex claims behaviour – a data quality issue with direct pricing implications.

------------------------------------------------------------------------

### 8.2 Glass Claims as a "Gateway Drug" to Larger Claims

**Technique justification:** Overlaid `geom_density()` comparing the claim frequency distributions for glass claimers vs non-claimers. The density overlay is preferred over a violin or boxplot because the focus is on *distributional differences* across the entire range – particularly whether glass claimers show a heavier right tail at claim frequencies above 1.0. The area-under-the-curve difference in the right tail is the key quantity; a median-focused chart would miss it.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-glass-gateway
#| fig-cap: "Claim frequency density: glass claimers vs non-claimers. Tests whether filing a glass claim is a leading indicator of future broader claims activity."
#| fig-width: 9
#| fig-height: 5

glass_data <- insurance_viz |>
  filter(total_exposure > 0, claim_frequency < 8) |>
  mutate(glass_claimer_label = if_else(had_glass_claim,
                                       "Filed glass claim",
                                       "No glass claim"))

ggplot(glass_data,
       aes(x      = claim_frequency,
           fill   = glass_claimer_label,
           colour = glass_claimer_label)) +
  geom_density(alpha = 0.35, linewidth = 0.9) +
  scale_fill_manual(
    values = c("Filed glass claim" = "#C0392B",
               "No glass claim"    = "#4B8BBE"),
    name   = NULL
  ) +
  scale_colour_manual(
    values = c("Filed glass claim" = "#922B21",
               "No glass claim"    = "#1A5276"),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(0, 5),
                     labels = number_format(accuracy = 0.1)) +
  labs(
    title    = "Claim Frequency: Glass Claimers vs Non-Claimers",
    subtitle = "Does a glass claim signal broader future claims activity?",
    x        = "Claim Frequency (claims per policy year)",
    y        = "Density"
  ) +
  theme(legend.position = "top")
```

## Code

```{r}
#| eval: false
#| label: fig-glass-gateway-code
#| fig-cap: "Claim frequency density: glass claimers vs non-claimers. Tests whether filing a glass claim is a leading indicator of future broader claims activity."
#| fig-width: 9
#| fig-height: 5

glass_data <- insurance_viz |>
  filter(total_exposure > 0, claim_frequency < 8) |>
  mutate(glass_claimer_label = if_else(had_glass_claim,
                                       "Filed glass claim",
                                       "No glass claim"))

ggplot(glass_data,
       aes(x      = claim_frequency,
           fill   = glass_claimer_label,
           colour = glass_claimer_label)) +
  geom_density(alpha = 0.35, linewidth = 0.9) +
  scale_fill_manual(
    values = c("Filed glass claim" = "#C0392B",
               "No glass claim"    = "#4B8BBE"),
    name   = NULL
  ) +
  scale_colour_manual(
    values = c("Filed glass claim" = "#922B21",
               "No glass claim"    = "#1A5276"),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(0, 5),
                     labels = number_format(accuracy = 0.1)) +
  labs(
    title    = "Claim Frequency: Glass Claimers vs Non-Claimers",
    subtitle = "Does a glass claim signal broader future claims activity?",
    x        = "Claim Frequency (claims per policy year)",
    y        = "Density"
  ) +
  theme(legend.position = "top")
```
:::

**Interpretation:** The density chart provides strong visual evidence for the "gateway drug" hypothesis. The non-claimer curve (blue) peaks sharply at zero claim frequency and falls away quickly – the vast majority of non-glass-claimers make no further claims. In stark contrast, the glass-claimer density curve (red) is bimodal: it has its primary mass around 1.0 claims per policy year and a secondary mode near 2.0, with a longer right tail extending beyond 3.0. This means policyholders who filed a glass claim in the same policy year **are concentrated at claim frequencies of 1–2 events**, not at zero. The red curve barely touches zero, confirming that glass claimers overwhelmingly have at least one other claim type in the same period. **This is a definitive confirmation of the gateway signal.** Glass claims do not occur in isolation – they co-occur with broader claims activity. The practical implication for underwriting is substantial: glass claim history should be incorporated into the bonus-malus calculation or used as a supplementary flag at renewal. A policyholder with a glass claim in the prior year should carry a moderate frequency loading on renewal, even if the glass claim itself is excluded from the formal bonus-malus calculation. Failing to do so systematically under-prices the renewal of these policyholders.

------------------------------------------------------------------------

### 8.3 New Business vs Portfolio: Who Is the Worse Risk, and Why?

**Technique justification:** `ggbetweenstats()` from **ggstatsplot** for a two-group comparison. This function combines a violin plot, boxplot, jittered individual data points, and an automated Wilcoxon test with effect size reporting into a single chart. For a binary grouping variable like `business_type`, this is the most analytically complete single-chart option: it answers not just "is there a difference?" but "how large is it and is it statistically significant?" without requiring separate charts.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-nb-vs-portfolio
#| fig-cap: "Loss ratio comparison: new business vs portfolio. ggbetweenstats automates the Wilcoxon test and reports effect size. The break-even reference line is added as a ggplot component."
#| fig-width: 8
#| fig-height: 6

ggbetweenstats(
  data                 = insurance_viz |>
    filter(loss_ratio_capped > 0, total_exposure > 0),
  x                    = business_type,
  y                    = loss_ratio_capped,
  type                 = "nonparametric",
  pairwise.comparisons = FALSE,
  title                = "Loss Ratio: New Business vs Portfolio",
  subtitle             = "Wilcoxon test; does the NB channel carry higher inherent risk?",
  xlab                 = "Business Type",
  ylab                 = "Loss Ratio",
  ggtheme              = theme_minimal(base_size = 11),
  ggplot.component     = list(
    geom_hline(yintercept = 1, linetype = "dashed",
               colour = "grey50", linewidth = 0.7),
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40")
    )
  )
)
```

## Code

```{r}
#| eval: false
#| label: fig-nb-vs-portfolio-code
#| fig-cap: "Loss ratio comparison: new business vs portfolio. ggbetweenstats automates the Wilcoxon test and reports effect size. The break-even reference line is added as a ggplot component."
#| fig-width: 8
#| fig-height: 6

ggbetweenstats(
  data                 = insurance_viz |>
    filter(loss_ratio_capped > 0, total_exposure > 0),
  x                    = business_type,
  y                    = loss_ratio_capped,
  type                 = "nonparametric",
  pairwise.comparisons = FALSE,
  title                = "Loss Ratio: New Business vs Portfolio",
  subtitle             = "Wilcoxon test; does the NB channel carry higher inherent risk?",
  xlab                 = "Business Type",
  ylab                 = "Loss Ratio",
  ggtheme              = theme_minimal(base_size = 11),
  ggplot.component     = list(
    geom_hline(yintercept = 1, linetype = "dashed",
               colour = "grey50", linewidth = 0.7),
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40")
    )
  )
)
```
:::

**Interpretation:** The effect size (rank-biserial correlation) reported by `ggbetweenstats` is as important as the p-value here: a statistically significant but negligibly small effect size would suggest the difference, while real, does not warrant a different pricing approach for new business. A large effect size, on the other hand, would confirm that the NB channel absorbs a materially worse risk mix and should carry a structural new business loading in the rating engine – separate from and in addition to any individual risk rating factors.

------------------------------------------------------------------------

## 9 Bivariate Analysis: Loyalty Drivers

### 9.1 Do Profitable Customers Cancel More Often?

**Technique justification:** A two-step approach. First, `ggbarstats()` tests whether cancellation rates differ by bonus score – confirming the structural pattern. Second, an overlaid `geom_density()` chart within each bonus score group shows whether the *cancelled* policies had better or worse loss ratios than those that stayed. The density overlay is essential for the second step: a simple mean comparison would miss the distributional story (whether the left tail – low loss ratio, profitable policies – is disproportionately represented in cancellations).

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-cancellation-bonus
#| fig-cap: "Cancellation rate by bonus score (ggbarstats). Tests whether Good-risk policyholders cancel at higher rates – the insurance retention paradox."
#| fig-width: 7
#| fig-height: 6

ggbarstats(
  data             = insurance_viz,
  x                = policy_status,
  y                = bonus_score,
  label            = "both",
  xlab             = "Bonus Score",
  ylab             = "Proportion",
  legend.title     = "Policy Status",
  title            = "Cancellation Rate by Bonus Score",
  subtitle         = "χ²: do Good-risk policyholders cancel at higher rates?",
  palette          = "Set2",
  ggtheme          = theme_minimal(base_size = 10),
  ggplot.component = list(
    scale_fill_manual(
      values = c("Active" = "#5C9E6E", "Cancelled" = "#C0392B"),
      name   = "Policy Status"
    ),
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
      axis.text.x   = element_text(size = 9)
    )
  )
)
```

## Code

```{r}
#| eval: false
#| label: fig-cancellation-bonus-code
#| fig-cap: "Cancellation rate by bonus score (ggbarstats). Tests whether Good-risk policyholders cancel at higher rates – the insurance retention paradox."
#| fig-width: 7
#| fig-height: 6

ggbarstats(
  data             = insurance_viz,
  x                = policy_status,
  y                = bonus_score,
  label            = "both",
  xlab             = "Bonus Score",
  ylab             = "Proportion",
  legend.title     = "Policy Status",
  title            = "Cancellation Rate by Bonus Score",
  subtitle         = "χ²: do Good-risk policyholders cancel at higher rates?",
  palette          = "Set2",
  ggtheme          = theme_minimal(base_size = 10),
  ggplot.component = list(
    scale_fill_manual(
      values = c("Active" = "#5C9E6E", "Cancelled" = "#C0392B"),
      name   = "Policy Status"
    ),
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
      axis.text.x   = element_text(size = 9)
    )
  )
)
```
:::

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-cancelled-loss-ratio
#| fig-cap: "Loss ratio density: active vs cancelled policies, faceted by bonus score. If the cancelled (red) density peaks left of active (blue), the insurer is losing its most profitable risks."
#| fig-width: 10
#| fig-height: 5

insurance_viz |>
  filter(loss_ratio_capped > 0, total_exposure > 0) |>
  ggplot(aes(x      = loss_ratio_capped,
             fill   = policy_status,
             colour = policy_status)) +
  geom_density(alpha = 0.35, linewidth = 0.9) +
  facet_wrap(~bonus_score, nrow = 1) +
  scale_fill_manual(
    values = c("Active" = "#4B8BBE", "Cancelled" = "#C0392B"),
    name   = "Policy Status"
  ) +
  scale_colour_manual(
    values = c("Active" = "#1A5276", "Cancelled" = "#922B21"),
    name   = "Policy Status"
  ) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.6) +
  scale_x_continuous(limits = c(0, 5)) +
  labs(
    title    = "Loss Ratio Distribution: Active vs Cancelled by Bonus Score",
    subtitle = "If cancelled (red) peaks left of active (blue): profitable customers are leaving",
    x        = "Loss Ratio (capped at 99th percentile)",
    y        = "Density"
  ) +
  theme(legend.position = "top")
```

## Code

```{r}
#| eval: false
#| label: fig-cancelled-loss-ratio-code
#| fig-cap: "Loss ratio density: active vs cancelled policies, faceted by bonus score. If the cancelled (red) density peaks left of active (blue), the insurer is losing its most profitable risks."
#| fig-width: 10
#| fig-height: 5

insurance_viz |>
  filter(loss_ratio_capped > 0, total_exposure > 0) |>
  ggplot(aes(x      = loss_ratio_capped,
             fill   = policy_status,
             colour = policy_status)) +
  geom_density(alpha = 0.35, linewidth = 0.9) +
  facet_wrap(~bonus_score, nrow = 1) +
  scale_fill_manual(
    values = c("Active" = "#4B8BBE", "Cancelled" = "#C0392B"),
    name   = "Policy Status"
  ) +
  scale_colour_manual(
    values = c("Active" = "#1A5276", "Cancelled" = "#922B21"),
    name   = "Policy Status"
  ) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.6) +
  scale_x_continuous(limits = c(0, 5)) +
  labs(
    title    = "Loss Ratio Distribution: Active vs Cancelled by Bonus Score",
    subtitle = "If cancelled (red) peaks left of active (blue): profitable customers are leaving",
    x        = "Loss Ratio (capped at 99th percentile)",
    y        = "Density"
  ) +
  theme(legend.position = "top")
```
:::

**Interpretation:** The two charts together confirm the retention paradox with empirical precision. From the first chart: Good-score policyholders cancel at a rate of **10%**, Neutral at **17%**, and Bad at **25%** – a monotonically increasing cancellation rate as risk quality *worsens*. At first glance this seems reassuring (riskier customers leave more). But the second chart reframes this entirely. Looking at the density overlays faceted by bonus score: in the **Good** panel, the cancelled (red) and active (blue) densities are very similar in shape and peak position, with the cancelled density showing a slightly heavier right tail – suggesting that among Good-score policyholders, those who *left* were not systematically the most profitable ones. In the **Neutral** and **Bad** panels, the cancelled policies show loss ratios that are broadly similar to or higher than the retained policies. **The combined picture tells a nuanced story:** the insurer is not clearly losing its best Good-score risks (the retention paradox is partial, not full), but it is also not successfully shedding its worst Bad-score risks at meaningful rates – 75% of Bad-score policyholders *stay*. The highest-risk cohort is the stickiest, likely because they struggle to obtain competitive quotes elsewhere. This has direct implications: the underwriter should aggressively price Bad-score renewals to not-renew or restructure them, rather than allowing price stickiness to retain a disproportionate concentration of poor-quality risks.

------------------------------------------------------------------------

### 9.2 Exposure Duration as a Hidden Quality Signal

**Technique justification:** `stat_density_ridges()` from **ggridges** with quantile shading, plotting loss ratio distributions across four exposure duration bands. The ridgeline format is chosen because comparing four distributional profiles simultaneously in a compact vertical layout is its primary use case. A boxplot of four groups would work, but would suppress the potentially bimodal or heavy-tailed shape within the short-duration bands that is key to the hypothesis.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-exposure-quality
#| fig-cap: "Loss ratio by exposure duration band. Tests whether short-duration policies carry systematically worse loss profiles – evidence of claim-then-cancel behaviour."
#| fig-width: 9
#| fig-height: 5

exposure_data <- insurance_viz |>
  filter(loss_ratio_capped > 0, total_exposure > 0) |>
  mutate(
    exposure_band = cut(total_exposure,
      breaks         = c(0, 0.35, 0.75, 0.95, Inf),
      labels         = c("Very short (<0.35yr)",
                         "Short (0.35–0.75yr)",
                         "Medium (0.75–0.95yr)",
                         "Full (>0.95yr)"),
      include.lowest = TRUE)
  )

ggplot(exposure_data,
       aes(x    = loss_ratio_capped,
           y    = exposure_band,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom           = "density_ridges_gradient",
    calc_ecdf      = TRUE,
    quantiles      = 4,
    quantile_lines = TRUE,
    scale          = 2.0,
    rel_min_height = 0.005,
    bandwidth      = 0.05
  ) +
  scale_fill_viridis_d(name = "Quartile", option = "C", alpha = 0.85) +
  scale_x_continuous(
    limits = c(0, 6),
    labels = number_format(accuracy = 0.1),
    expand = c(0, 0)
  ) +
  scale_y_discrete(expand = expansion(add = c(0.3, 2.5))) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  labs(
    title    = "Loss Ratio Distribution by Exposure Duration",
    subtitle = "Ridgeline with quartile shading; dashed line = break-even (LR = 1)",
    x        = "Loss Ratio (capped at 99th percentile)",
    y        = "Exposure\nDuration Band"
  ) +
  theme_ridges(font_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
    legend.position = "right"
  )
```

## Code

```{r}
#| eval: false
#| label: fig-exposure-quality-code
#| fig-cap: "Loss ratio by exposure duration band. Tests whether short-duration policies carry systematically worse loss profiles – evidence of claim-then-cancel behaviour."
#| fig-width: 9
#| fig-height: 5

exposure_data <- insurance_viz |>
  filter(loss_ratio_capped > 0, total_exposure > 0) |>
  mutate(
    exposure_band = cut(total_exposure,
      breaks         = c(0, 0.35, 0.75, 0.95, Inf),
      labels         = c("Very short (<0.35yr)",
                         "Short (0.35–0.75yr)",
                         "Medium (0.75–0.95yr)",
                         "Full (>0.95yr)"),
      include.lowest = TRUE)
  )

ggplot(exposure_data,
       aes(x    = loss_ratio_capped,
           y    = exposure_band,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom           = "density_ridges_gradient",
    calc_ecdf      = TRUE,
    quantiles      = 4,
    quantile_lines = TRUE,
    scale          = 2.0,
    rel_min_height = 0.005,
    bandwidth      = 0.05
  ) +
  scale_fill_viridis_d(name = "Quartile", option = "C", alpha = 0.85) +
  scale_x_continuous(
    limits = c(0, 6),
    labels = number_format(accuracy = 0.1),
    expand = c(0, 0)
  ) +
  scale_y_discrete(expand = expansion(add = c(0.3, 2.5))) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  labs(
    title    = "Loss Ratio Distribution by Exposure Duration",
    subtitle = "Ridgeline with quartile shading; dashed line = break-even (LR = 1)",
    x        = "Loss Ratio (capped at 99th percentile)",
    y        = "Exposure\nDuration Band"
  ) +
  theme_ridges(font_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
    legend.position = "right"
  )
```
:::

**Interpretation:** The ridgeline chart reveals a striking and unambiguous pattern. The **Full (\>0.95yr)** band shows the most compact loss ratio distribution, concentrated just below the break-even line of 1.0, with the narrowest spread. These are the portfolio's most stable, predictable policies. As exposure duration decreases, the distributions progressively widen and shift rightward. The **Very short (\<0.35yr)** band has the heaviest right tail, extending well beyond 3.0 and with a substantial proportion of policies above the break-even line. Crucially, even the **median** loss ratio for short-duration policies sits to the right of the full-duration median confirming that short-duration policies are not just more volatile but systematically less profitable. The pattern is consistent with short-duration policies carrying a higher loss burden relative to the premium collected. One plausible explanation is that policyholders who experience a claim early may cancel before the policy year ends, producing high incurred costs relative to partial-year premium. However, this dataset does not contain the within-year timing of claims and cancellations, so claim-then-cancel behaviour cannot be directly confirmed from the data. It remains a hypothesis consistent with the observed pattern. What can be stated from the data is that very short duration policies show a systematically worse loss ratio distribution than full-duration policies, and this pattern is visible and consistent across all four exposure bands in the ridgeline chart.

------------------------------------------------------------------------

## 10 Bivariate Analysis: Geographic Drivers

### 10.1 Loss Ratio by Municipality Type and Circulation Area

**Technique justification:** `stat_pointinterval()` from **ggdist** with simultaneous 95% and 99% confidence intervals. A boxplot or grouped bar chart of summary medians is explicitly avoided because it conveys false precision for the Islands segment, which has a small policy count (as confirmed in Section 4.3). `stat_pointinterval()` makes the uncertainty visible: wide intervals signal that an estimate is unreliable, which is an analytical finding as important as the point estimate itself. This is the only chart in the report where the width of the interval – not the position of the point – is the primary message for at least one segment.

::: panel-tabset
## Plot

```{r}
#| echo: false
#| label: fig-lr-geography
#| fig-cap: "Loss ratio by geographic segment. Median + 95%/99% CIs. Wide island confidence intervals are the finding: the island loss ratio estimate is highly uncertain, not just high."
#| fig-width: 10
#| fig-height: 5

geo_data <- insurance_viz |>
  filter(loss_ratio_capped > 0, total_exposure > 0) |>
  mutate(
    geo_segment = paste0(municipality_type, "\n(", circulation_area, ")"),
    geo_segment = fct_reorder(geo_segment, loss_ratio_capped, median)
  )

ggplot(geo_data,
       aes(x = geo_segment, y = loss_ratio_capped)) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 3.5,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    name   = "Confidence\nInterval",
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95% CI", "99% CI")
  ) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  annotate("text", x = 0.6, y = 1.04,
           label = "Break-even (LR = 1)",
           hjust = 0, size = 3, colour = "grey50") +
  labs(
    title    = "Loss Ratio by Geographic Segment",
    subtitle = "Ordered by median; wide CIs on island segments reflect small sample volumes",
    x        = "Geographic Segment\n(Municipality × Circulation Area)",
    y        = "Loss\nRatio"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10))
```

## Code

```{r}
#| eval: false
#| label: fig-lr-geography-code
#| fig-cap: "Loss ratio by geographic segment. Median + 95%/99% CIs. Wide island confidence intervals are the finding: the island loss ratio estimate is highly uncertain, not just high."
#| fig-width: 10
#| fig-height: 5

geo_data <- insurance_viz |>
  filter(loss_ratio_capped > 0, total_exposure > 0) |>
  mutate(
    geo_segment = paste0(municipality_type, "\n(", circulation_area, ")"),
    geo_segment = fct_reorder(geo_segment, loss_ratio_capped, median)
  )

ggplot(geo_data,
       aes(x = geo_segment, y = loss_ratio_capped)) +
  stat_pointinterval(
    .width     = c(0.95, 0.99),
    point_size = 3.5,
    aes(colour = factor(after_stat(.width)))
  ) +
  scale_colour_manual(
    name   = "Confidence\nInterval",
    values = c("0.95" = "#4B8BBE", "0.99" = "#E07B39"),
    labels = c("95% CI", "99% CI")
  ) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  annotate("text", x = 0.6, y = 1.04,
           label = "Break-even (LR = 1)",
           hjust = 0, size = 3, colour = "grey50") +
  labs(
    title    = "Loss Ratio by Geographic Segment",
    subtitle = "Ordered by median; wide CIs on island segments reflect small sample volumes",
    x        = "Geographic Segment\n(Municipality × Circulation Area)",
    y        = "Loss\nRatio"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10))
```
:::

**Interpretation:** Urban segments consistently show higher median loss ratios than rural equivalents within each municipality type, confirming circulation area as a genuine geographic risk discriminator. The island segments display substantially wider confidence intervals than inland segments regardless of their median values. This means the insurer does not have enough data to set island pricing with statistical confidence. The appropriate response is either to grow the island book to tighten the interval, or to apply an explicit uncertainty loading over and above any expected value estimate. Using only the observed median loss ratio for island pricing without accounting for the wide confidence intervals would treat an imprecise estimate as though it were a reliable figure. The confidence interval widths shown in this chart indicate the estimate is not yet statistically stable at these policy volumes.

------------------------------------------------------------------------

## 11 Summary of Key Findings and Underwriting Implications

This section distils the most actionable findings from the visual analytics investigation into a concise reference for the motor insurance underwriter. Findings are ranked by commercial priority.

### 11.1 Portfolio Profitability: Critical Findings

**Finding 1 – COMP_N carries the most adverse risk profile of all product types.** Comprehensive cover without excess (COMP_N) carries the highest proportion of Bad bonus score policyholders (3.5% vs 1.0%–2.0% for other types) and its median loss ratio sits just above break-even — the combination of adverse selection in its risk intake and a median position above LR = 1 makes it the product type with the most concerning risk-profitability combination. Note that Third-Party shows the highest median loss ratio of all product types, though this reflects its minimal coverage structure. Observed pattern: COMP_N has the highest Bad bonus concentration of all product types and its median loss ratio sits above LR = 1. Third-Party has the highest median loss ratio overall.

**Finding 2 – Bonus score is associated with significantly different loss ratio distributions, but the direction is counterintuitive.** All three pairwise Wilcoxon tests are statistically significant (p < 0.0001 for Good vs Neutral and Good vs Bad; p < 0.01 for Neutral vs Bad). However, Good-score policyholders show the highest median loss ratio (~2.0) and widest distribution, while Neutral and Bad groups show lower medians (~1.5) and more compressed distributions. This counterintuitive finding warrants further investigation — the bonus score classification may not be functioning as a simple risk discriminator in this portfolio, or the Good group may systematically include higher-value policies generating larger absolute losses. Observed pattern: the half-eye chart confirms all three groups have statistically different distributions, with Good showing the highest median and widest spread.

**Finding 3 – Short-duration policies show systematically worse loss ratio distributions than full-duration policies.** Very short (<0.35yr) and short (0.35–0.75yr) duration policies show higher median loss ratios and heavier right tails than full-year policies. The pattern is consistent with short-duration policies carrying a higher loss burden relative to the premium collected. The within-year timing of claims and cancellations is not available in this dataset, so claim-then-cancel behaviour remains a hypothesis consistent with the observed pattern rather than a confirmed finding. Observed pattern: the ridgeline chart shows a progressive rightward shift and widening of distributions as exposure duration decreases, consistent and visible across all four bands.

**Finding 4 – New business carries materially worse risk quality than portfolio renewals.** NB policyholders carry 1.6% Bad bonus proportion vs 0.8% for portfolio renewals — a 100% relative excess. The Neutral proportion gap (2.6% vs 1.3%) is similarly large. Observed pattern: the chi-square test confirms new business carries a significantly higher proportion of Bad and Neutral bonus score policyholders than portfolio renewals (p < 0.001). The magnitude of this difference is visible in the stacked proportion charts.

### 11.2 Claims Volatility: Key Drivers

**Finding 5 – Glass claims co-occur with broader claims activity in the same policy year.** Policyholders who file glass claims show bimodal claim frequency distributions centred at 1.0 and 2.0 events per year — not near zero like non-glass-claimers. The red density curve barely touches zero, confirming glass claimers overwhelmingly have at least one other claim type in the same period. Observed pattern: the density overlay chart shows glass claimers have a bimodal frequency distribution centred at 1.0 and 2.0 events per year, compared to non-claimers concentrated near zero.

**Finding 6 – Gasoline vehicles show marginally higher median claim severity than diesel vehicles (Wilcoxon p = 0.04).** The chart shows Gasoline's median line sitting slightly above Diesel's on the log scale. Both distributions span a similar range with bulk density concentrated between €300 and €1,000. The Gasoline half-eye extends slightly further rightward above €1,000. The mechanism cannot be determined from this dataset alone. Observed pattern: the Wilcoxon test confirms a statistically significant difference in median claim severity (p = 0.04), with Gasoline showing marginally higher severity than Diesel. The half-eye shapes are similar in form, indicating a location shift rather than a tail effect.

**Finding 7 – Claim severity does not show a clear systematic relationship with vehicle age across most of the distribution. **  The GAM smoother on the hexbin chart is largely flat across the 0–50 year vehicle age range. A marginal upward trend is visible for vehicles aged 50+ years, but confidence intervals at this extreme are wide due to sparse data. The claimed U-shaped pattern (highest for new, declining, then uptick for old) is not clearly supported by the rendered GAM line. Observed pattern: the hexbin density is broadly distributed across vehicle ages 5–50 years without a clear directional trend in the GAM smoother. A marginal late-age uptick is visible but uncertainty is high.

### 11.3 Segmentation and Adverse Selection

**Finding 8 – Luxury brand tier carries a statistically confirmed but small adverse selection signal (χ²=281.9, p < 0.001, small Cramér's V). ** Luxury vehicles carry 2.0% Bad bonus proportion vs 1.2%–1.3% for other tiers. The severity 99% CI extends further upward for Luxury than other tiers, suggesting greater variability in extreme claim costs. Whether this warrants a separate brand tier loading requires further multivariate analysis. This dataset identifies the association but cannot determine independent predictive power once other variables are controlled for. Observed pattern: chi-square confirms significant association with small effect size. Severity 99% CI is wider for Luxury, suggesting greater variability in extreme claims.

**Finding 9 – Power-to-weight ratio shows no detectable association with claim frequency in this portfolio. ** GAM smoothers are flat across all brand tiers with near-zero Spearman ρ values. Observed pattern: flat GAM curves and near-zero ρ values across all three brand tiers indicate no detectable association between power-to-weight ratio and claim frequency in this dataset.

**Finding 10 – Bad-score policyholders are the stickiest cohort: 75% remain active vs 90% of Good-score policies. ** The insurer's natural attrition process does not effectively shed its worst risks. 75% of Bad-score policyholders stay, likely because they struggle to obtain competitive alternatives elsewhere. Observed pattern: Bad-score policyholders have a 25% cancellation rate versus 10% for Good-score policyholders. Despite the higher cancellation rate, 75% of Bad-score policyholders remain active, the largest absolute retention of any bonus group.

### 11.4 Geographic Risk

**Finding 11 – Urban policies show higher loss ratios than rural; island segment estimates are statistically unreliable.** Urban segments show consistently higher median loss ratios than rural equivalents within each municipality type. Island segments show dramatically wider confidence intervals than inland segments, reflecting smaller policy volumes. Using only the observed median loss ratio for island pricing without accounting for the wide confidence intervals would treat an imprecise estimate as a reliable figure. Observed pattern: urban segments consistently show higher median loss ratios than rural equivalents across all municipality types. Island segments show substantially wider confidence intervals than inland segments.

------------------------------------------------------------------------

-   [R for Visual Analytics](https://r4va.netlify.app/)
-   [R for Data Science (2nd ed.)](https://r4ds.hadley.nz/)
-   [ggplot2 Extensions Gallery](https://exts.ggplot2.tidyverse.org/gallery/)
-   [ggridges documentation](https://wilkelab.org/ggridges/)
-   [ggdist: Visualizations of Distributions and Uncertainty](https://mjskay.github.io/ggdist/)
-   [ggstatsplot: ggplot2-Based Plots with Statistical Details](https://indrajeetpatil.github.io/ggstatsplot/)
-   Wilke, C. O. (2019). *Fundamentals of Data Visualisation*. O'Reilly Media. <https://clauswilke.com/dataviz/>
-   Klugman, S. A., Panjer, H. H., & Willmot, G. E. (2019). *Loss Models: From Data to Decisions* (5th ed.). John Wiley & Sons.
-   Frees, E. W. et al. *Loss Data Analytics*, Chapter 3: Modeling Loss Severity. <https://openacttexts.github.io/Loss-Data-Analytics/ChapSeverity.html>
-   Venegas-Martínez, F. et al. (2014). *A Lognormal Model for Insurance Claims Data*. ResearchGate. <https://www.researchgate.net/publication/252507857>
-   Meng, S., Gao, Y., & Huang, Y. (2022). Actuarial intelligence in auto insurance: Claim frequency modeling with driving behavior features and improved boosted trees. *Insurance: Mathematics and Economics*, 106, 115–127. <https://doi.org/10.1016/j.insmatheco.2022.06.001>
-   Angelo, M. (2024). Analysis of Claim Frequency and Severity in Auto Insurance Using Generalized Linear Models in Philippines. *Journal of Statistics and Actuarial Research*, 8(3), 34–46. <https://doi.org/10.47604/jsar.2900>
-   Association of British Insurers. (2009). *ABI Research Paper No. 12: Insurance and Age-Based Differentiation*. London: ABI. <https://media.crai.com/sites/default/files/publications/insurance-and-age-based-differentiation.pdf>
-   Cappelletti, D. (2023). Older Drivers: Road Safety and Automobile Insurance. *SOA General Insurance Insights*, June 2023. <https://www.soa.org/publications/gi-insights/2023/06/gii-2023-06-cappelletti-2/>
-   Mendeley Data. Motor Vehicle Insurance Portfolio Dataset. <https://data.mendeley.com/datasets/sw4jmdb2sm/1>

**\~\~\~ End of Take-home Exercise 01 \~\~\~**
