---
title: "Visual Analytics for Motor Vehicle Insurance"
subtitle: "ISSS608 Visual Analytics Individual Assignment"
author: "Ang Ming Kwan"
date: "May 15, 2026"
date-modified: "last-modified"
format:
  revealjs:
    theme: default
    slide-number: false
    transition: fade
    transition-speed: fast
    progress: true
    scrollable: false
    smaller: false
    margin: 0.05
    include-after-body:
      text: |
        <style>
          /* ── Global size increase ~22% ─────────────────────────── */
          .reveal { font-size: 22px !important; }
          .reveal h2 {
            font-size: 1.35em !important;
            margin-bottom: 0.25em !important;
            line-height: 1.2 !important;
          }
          .reveal h3 {
            font-size: 1.05em !important;
            margin-bottom: 0.2em !important;
          }
          .reveal p, .reveal li {
            font-size: 1em !important;
            line-height: 1.4 !important;
            margin-bottom: 0.4em !important;
          }
          .reveal .slides section {
            padding-top: 0.1em !important;
          }
          /* ── Table cells ───────────────────────────────────────── */
          .reveal table td, .reveal table th {
            font-size: 0.9em !important;
            padding: 4px 8px !important;
            line-height: 1.35 !important;
          }
          /* ── Blockquote ────────────────────────────────────────── */
          .reveal blockquote {
            font-size: 0.88em !important;
            padding: 5px 12px !important;
            margin: 5px 0 !important;
          }
          /* ── Inline code ───────────────────────────────────────── */
          .reveal code { font-size: 0.88em !important; }
          /* ── Slide number badge ────────────────────────────────── */
          .slide-number-custom {
            position: fixed; bottom: 14px; right: 18px;
            font-size: 12px; font-family: var(--r-main-font, sans-serif);
            color: #555; background: rgba(255,255,255,0.85);
            padding: 2px 7px; border-radius: 4px;
            z-index: 9999; pointer-events: none;
          }
        </style>
        <div class="slide-number-custom" id="custom-slide-num">slide 1 / 15</div>
        <script>
          Reveal.on('slidechanged', function(event) {
            var total = Reveal.getTotalSlides();
            var current = Reveal.getState().indexh + 1;
            document.getElementById('custom-slide-num').textContent =
              'slide ' + current + ' / ' + total;
          });
          Reveal.on('ready', function() {
            var total = Reveal.getTotalSlides();
            var current = Reveal.getState().indexh + 1;
            document.getElementById('custom-slide-num').textContent =
              'slide ' + current + ' / ' + total;
          });
        </script>
execute:
  eval: true
  echo: false
  warning: false
  message: false
  freeze: false
---

```{r}
#| label: pres-setup
#| results: hide
pacman::p_load(
  tidyverse, readr, ggdist, ggridges,
  ggstatsplot, ggpubr, patchwork,
  scales, viridis, knitr, kableExtra
)

insurance <- read_delim(
  "data/Dataset_of_motor_insurance_portfolio.csv",
  delim = ";", show_col_types = FALSE
) |>
  mutate(
    policy_type = factor(policy_type,
      levels = c("TP","TPG","CC","COMP_E","COMP_N"),
      labels = c("Third-Party","TP + Glass","Combined Cover",
                 "Comp. (Excess)","Comp. (No Excess)")),
    policy_status = factor(policy_status,
      levels = c("A","C"), labels = c("Active","Cancelled")),
    business_type = factor(business_type,
      levels = c("NB","P"), labels = c("New Business","Portfolio")),
    bonus_score = factor(bonus_score,
      levels = c("G","N","B"),
      labels = c("Good","Neutral","Bad"), ordered = TRUE),
    fuel_type = factor(fuel_type,
      levels = c("D","G"), labels = c("Diesel","Gasoline")),
    municipality_type = factor(municipality_type,
      levels = c("I","C","IS"), labels = c("Inland","Coastal","Islands")),
    circulation_area = factor(circulation_area,
      levels = c("U","R"), labels = c("Urban","Rural")),
    loss_ratio      = total_incurred / total_premium,
    claim_frequency = total_claims / total_exposure,
    claim_severity  = if_else(total_claims > 0,
                              total_incurred / total_claims, NA_real_),
    pure_premium    = total_incurred / total_exposure,
    years_licensed  = year - age_driving_licence,
    licence_age     = driver_age - years_licensed,
    driver_age_band = cut(driver_age, c(17,25,35,45,55,65,Inf),
      labels = c("18-25","26-35","36-45","46-55","56-65","65+"),
      right = TRUE, include.lowest = TRUE),
    vehicle_age_band = cut(vehicle_age,
      breaks = c(-Inf,3,7,12,18,25,Inf),
      labels = c("0-3","4-7","8-12","13-18","19-25","25+"), right = TRUE),
    brand_tier = case_when(
      vehicle_brand %in% c("BMW","MERCEDES","AUDI","LEXUS","PORSCHE",
        "JAGUAR","LAND ROVER","VOLVO","ALFA ROMEO") ~ "Luxury",
      vehicle_brand %in% c("VOLKSWAGEN","TOYOTA","FORD","OPEL","SEAT",
        "PEUGEOT","CITROEN","NISSAN","HYUNDAI","KIA",
        "FIAT","HONDA","MAZDA","SKODA","MITSUBISHI") ~ "Upper-mid",
      is.na(vehicle_brand) ~ NA_character_,
      TRUE ~ "Budget/Comm."),
    brand_tier = factor(brand_tier, levels = c("Luxury","Upper-mid","Budget/Comm.")),
    claim_breadth = rowSums(across(
      c(liability_claims,property_claims,theft_claims,
        fire_claims,glass_claims,legal_protection_claims,occupants_claims),
      \(x) x > 0)),
    had_glass_claim = glass_claims > 0
  )

lr_cap  <- quantile(insurance$loss_ratio,    0.99, na.rm = TRUE)
sev_cap <- quantile(insurance$claim_severity, 0.99, na.rm = TRUE)

insurance_viz <- insurance |>
  mutate(
    loss_ratio_capped = pmin(loss_ratio,    lr_cap),
    severity_capped   = pmin(claim_severity, sev_cap)
  )

bonus_pal <- c("Good" = "#27ae60", "Neutral" = "#e67e22", "Bad" = "#c0392b")

thm <- theme_minimal(base_size = 10) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5, size = 11),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey40", size = 8.5),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 8)
  )
theme_set(thm)
```

------------------------------------------------------------------------

## Slide 1 \| Executive Summary

::::::: columns
::::: {.column width="58%"}
::: {style="font-size:0.90em; line-height:1.35;"}
A visual analytics investigation of **354,140 motor insurance policies** from a Spanish non-life insurer (2022–2024) identifies structural profitability risks across product, demographic, behavioural, and geographic dimensions.

**Four critical findings demand underwriting attention:**

🔴 **COMP_N** (Comprehensive, No Excess): most loss-making product; 3.5% Bad-bonus policyholders confirms adverse selection; immediate actuarial review required

🔴 **Bonus score**: statistically validated predictor; loss ratio increases monotonically Good→Bad (all Wilcoxon p \< 0.05); Bad-score policyholders drive disproportionate large-loss tail exposure

🟠 **Short-duration policies**: systematically higher loss ratios confirm claim-then-cancel behaviour; structural premium inadequacy

🟠 **Glass claims are gateway events**: bimodal frequency at 1–2 events/year confirms co-occurrence with broader claims; validates glass history as a renewal risk flag
:::

::: {style="font-size:0.90em; margin-top:4px; padding:5px 10px; background:#fef9e7; border-left:3px solid #e67e22; border-radius:3px;"}
Also: New Businesses carries 2× adverse risk vs portfolio renewals +++ Bad-score policyholders: 75% retention +++ Urban \> Rural loss ratios +++ Island pricing statistically unreliable
:::
:::::

::: {.column width="42%"}
```{r}
#| label: exec-kpi-bar
#| fig-height: 3.7
#| fig-width: 4.2
kpi_df <- tibble(
  metric   = c("Policies","Active","Cancelled","With Claims","Zero Claims"),
  value    = c(354140,
               sum(insurance_viz$policy_status=="Active"),
               sum(insurance_viz$policy_status=="Cancelled"),
               sum(insurance_viz$total_claims > 0),
               sum(insurance_viz$total_claims == 0)),
  category = c("Total","Status","Status","Claims","Claims")
) |>
  mutate(pct   = value / 354140 * 100,
         label = paste0(comma(value)," (",round(pct,1),"%)"),
         metric = fct_rev(fct_inorder(metric)))

ggplot(kpi_df, aes(x = metric, y = pct, fill = category)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = label), hjust = -0.06, size = 2.8) +
  scale_fill_manual(values = c("Total"="#2c3e50","Status"="#2980b9","Claims"="#c0392b")) +
  scale_y_continuous(limits = c(0,120), labels = function(x) paste0(x,"%")) +
  coord_flip() +
  labs(title = "Portfolio at a Glance",
       subtitle = "354,140 policies | 2022–2024",
       x = NULL, y = "% of portfolio") +
  theme(plot.background = element_rect(fill = "white", colour = NA),
        axis.text.y = element_text(size = 9))
```
:::
:::::::

------------------------------------------------------------------------

## Slide 2 \| Data Preparation

:::::::: columns
::::: {.column width="55%"}
::: {style="font-size:0.90em; line-height:1.35;"}
**Import:** Explicit semicolon delimiter via `read_delim()`: prevents silent mis-parsing of the 47-column, 354,140-row file.

**Quality checks:** Zero duplicates. Four variables with missing values: handled **per-analysis** (not global imputation), keeping each decision transparent.

**Categorical recoding:** Raw codes (e.g. `"G"`, `"NB"`, `"U"`) replaced with labels. `bonus_score` converted to ordered factor: enabling correct ordering in Wilcoxon tests and ridgeline plots.

**Outlier capping:** 99th percentile cap on loss ratio and severity *for visualisation only*: all statistical tests use uncapped values.

**Binning:** Driver/vehicle age binned into actuarial bands. Vehicle brands classified into three tiers. `claim_breadth` (number of distinct claim lines per policy) derived as a behavioural signal variable.
:::

::: {style="font-size:0.90em; margin-top:5px; padding:5px 10px; background:#eaf3fb; border-left:3px solid #2980b9; border-radius:3px;"}
**Principle:** Every wrangling decision is documented at the point of use and not globally applied. This is to make the audit trail always traceable.
:::
:::::

:::: {.column width="45%"}
```{r}
#| label: slide2-missing
#| fig-height: 2.6
#| fig-width: 4.4
missing_df <- tibble(
  variable = c("fuel_type","vehicle_value","vehicle_age","age_driving_licence"),
  missing  = c(1287, 513, 2, 2),
  pct      = c(1287, 513, 2, 2) / 354140 * 100
) |>
  mutate(variable = fct_reorder(variable, missing), high = pct > 0.3)

ggplot(missing_df, aes(x = variable, y = missing, fill = high)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(missing)," (",round(pct,2),"%)")),
            hjust = -0.08, size = 2.8) +
  scale_fill_manual(values = c("FALSE"="#95a5a6","TRUE"="#e74c3c")) +
  scale_y_continuous(limits = c(0,1800), labels = comma) +
  coord_flip() +
  labs(title = "Missing Values by Variable",
       subtitle = "Red = materially missing (>0.3%)",
       x = NULL, y = "Count")
```

::: {style="font-size:0.90em; margin-top:6px; padding:5px 10px; background:#fef9e7; border-left:3px solid #e67e22; border-radius:3px;"}
`fuel_type` (0.36%) and `vehicle_value` (0.14%) are the only meaningful gaps. Impact on downstream analyses is minimal; excluded only in analyses that use those variables.
:::
::::
::::::::

------------------------------------------------------------------------

## Slide 3 \| Derived KPIs

::::::: columns
::::: {.column width="50%"}
:::: {style="font-size:0.90em; line-height:1.35;"}
**Why raw data is insufficient:** The dataset has premiums, claims, incurred costs, and exposure but none of the actuarial metrics needed to assess profitability or volatility. Six KPIs were derived as the analytical backbone.

**Exposure-adjustment is non-negotiable:** A 3-month policy and a 12-month policy generate the same raw claim count for zero incidents. `total_exposure` (policy years) normalises all KPIs. Without it, short-duration segments appear artificially safer.

**Demographic variables tested:** `years_licensed` and `licence_age` were derived to test whether driving experience predicts risk better than age.

::: {style="padding:5px 10px; background:#fdf2f8; border-left:3px solid #c0392b; border-radius:3px;"}
⚠️ **Data quality finding:** Both variables showed near-zero Spearman ρ with claim frequency because `age_driving_licence` records the *calendar year* of licence issue, and not years of experience. `years_licensed` is invalid as derived and requires re-specification before use in any model.
:::
::::
:::::

::: {.column width="50%"}
```{r}
#| label: slide3-kpi-table
kpi_tbl <- tibble(
  KPI = c("loss_ratio","claim_frequency","claim_severity",
          "pure_premium","years_licensed","licence_age"),
  Formula = c("incurred / premium",
              "claims / exposure",
              "incurred / claims",
              "incurred / exposure",
              "year − lic_year",
              "age − years_lic"),
  Meaning = c(
    "LR > 1 = underwriting loss",
    "Claims per policy year",
    "Avg cost per claim event",
    "Risk cost per policy year",
    "Experience proxy ⚠️",
    "Age at first licence")
)
knitr::kable(kpi_tbl, format = "html", escape = FALSE,
             align = c("l","l","l"),
             col.names = c("KPI","Formula","Interpretation")) |>
  kable_styling(bootstrap_options = c("striped","condensed"),
                full_width = TRUE, font_size = 11.5) |>
  column_spec(1, bold = TRUE, color = "#1a5276") |>
  row_spec(c(1,2), background = "#fef9e7") |>
  row_spec(c(3,4), background = "#eafaf1") |>
  row_spec(c(5,6), background = "#fdf2f8")
```
:::
:::::::

------------------------------------------------------------------------

## Slide 4 \| Analyses Overview: Univariate

::: {style="font-size:0.70em;"}
| \# | Analysis | Chart Type | Rationale |
|:-----------------|:-----------------|:-----------------|:-----------------|
| 4.1 | Portfolio composition by policy type | `ggbarstats()` + Pearson χ² | Plain bar only answers *how many*. `ggbarstats()` tests whether bonus score composition differs significantly across types |
| 4.2 | Bonus score by business type | `ggbarstats()` + Pearson χ² | Tests whether new business carries materially worse risk profile than portfolio renewals |
| 4.3 | Geographic distribution | Stacked 100% bar + count labels | Proportions **and** absolute volumes both matter; pure proportions hide the Islands segment's thin \~24,000-policy base |
| 4.4 | Distribution of vehicle age | Histogram + density overlay | Right-skewed integer variable; density curve smooths the empirical shape |
| 4.5 | Distribution of vehicle value | Histogram + density (log₁₀) | **Mandatory log scale**; two orders of magnitude spread compresses the modal range into invisibility on linear scale |
| 4.6 | Distribution of driver age | Histogram + density + reference lines | Reference lines at 25 and 65 make actuarial risk thresholds immediately visible |
:::

::: {style="font-size:0.70em; margin-top:4px; padding:5px 12px; background:#eaf4fc; border-left:3px solid #2980b9; border-radius:3px;"}
**Guiding principle:** Chart type is determined by the question, not convention. `ggbarstats()` replaces plain bars for categorical variables because the chi-square test is the primary analytical question. Log scale for vehicle value is a statistical necessity, not a stylistic choice.
:::

------------------------------------------------------------------------

## Slide 5 \| Analyses Overview: Bivariate

::: {style="font-size:0.6em;"}
| Domain | Analysis | Chart | Rationale |
|:-----------------|:-----------------|:-----------------|:-----------------|
| Conventional | Loss ratio by policy type | Violin + boxplot + Kruskal-Wallis | Reveals multi-modality; non-parametric for skewed data |
| Conventional | Loss ratio by bonus score | `stat_halfeye` + Wilcoxon pairwise | Tests monotonic Good→Bad ordering statistically |
| Conventional | Severity by fuel type | `stat_halfeye` + Wilcoxon (log) | Log scale mandatory; distinguishes location shift vs tail effect |
| Conventional | Severity by vehicle age | `geom_hex` + GAM | Hexbin resolves 354K-row overplotting; GAM captures non-linearity |
| Demographic | Loss ratio by age bands | Ridgelines + quartile shading | 6-group distributional shape comparison, not just medians |
| Demographic | Driver age × vehicle age | Heatmap (6×6) | Only compact format for 36-cell interaction matrix |
| Demographic | Driver age vs years licensed | Dual GAM + Spearman ρ | Side-by-side comparison of two candidate risk predictors |
| Brand/Status | Brand tier: frequency vs severity | Side-by-side `stat_pointinterval` | Decomposes dimensions conflated in a single loss ratio |
| Brand/Status | Bonus score by brand tier | `ggbarstats()` + χ² | Tests whether risk profile differs by brand tier |
| Brand/Status | Power-to-weight vs freq | `geom_hex` + GAM (faceted) | Tests performance-proxy signal across tiers |
| Brand/Status | Repair cost intensity | Ranked dotplot + CI | CI width communicates estimation confidence |
| Behavioural | Serial claimer breadth | Heatmap | Exposes anomalous multi-line claimers by bonus score |
| Behavioural | Glass claims gateway | Overlaid density | Right-tail differences confirm gateway signal |
| Behavioural | New business vs portfolio | `ggbetweenstats` | Unified violin + significance test + effect size |
| Loyalty | Cancellation paradox | `ggbarstats` + density | Two-step: cancellation rate, then loss ratio of leavers |
| Loyalty | Exposure duration | Ridgelines (4 bands) | Shape comparison across short vs full duration |
| Geographic | Loss ratio by geography | Point-interval 95%+99% CI | CI width is the primary finding for island segments |
:::

------------------------------------------------------------------------

## Slide 6 \| Univariate: Portfolio Composition

::::: columns
::: {.column width="50%"}
```{r}
#| label: slide6-policy-type
#| fig-height: 2.9
#| fig-width: 5.0
pt_data <- insurance_viz |>
  count(policy_type, bonus_score) |>
  group_by(policy_type) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(bonus_score = factor(bonus_score, c("Good","Neutral","Bad")))

ggplot(pt_data,
       aes(x = fct_rev(policy_type), y = pct, fill = bonus_score)) +
  geom_col(position = "stack", width = 0.65, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = ifelse(pct > 1.5, paste0(round(pct,1),"%"), "")),
            position = position_stack(vjust = 0.5),
            size = 2.6, colour = "white", fontface = "bold") +
  scale_fill_manual(values = bonus_pal, name = "Bonus") +
  scale_x_discrete(labels = function(x) str_wrap(x, 14)) +
  scale_y_continuous(labels = function(x) paste0(x,"%")) +
  coord_flip() +
  labs(title = "Policy Type × Bonus Score",
       subtitle = "χ² p < 0.001",
       x = NULL, y = "Proportion (%)") +
  theme(legend.position = "top")
```
:::

::: {.column width="50%"}
```{r}
#| label: slide6-bonus-business
#| fig-height: 2.9
#| fig-width: 4.8
nb_data <- insurance_viz |>
  count(business_type, bonus_score) |>
  group_by(business_type) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(bonus_score = factor(bonus_score, c("Good","Neutral","Bad")))

ggplot(nb_data,
       aes(x = business_type, y = pct, fill = bonus_score)) +
  geom_col(position = "stack", width = 0.5, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = paste0(round(pct,1),"%")),
            position = position_stack(vjust = 0.5),
            size = 2.8, colour = "white", fontface = "bold") +
  scale_fill_manual(values = bonus_pal, name = "Bonus") +
  scale_y_continuous(labels = function(x) paste0(x,"%")) +
  labs(title = "Bonus Score: NB vs Portfolio",
       subtitle = "NB: 1.6% Bad vs 0.8% Portfolio",
       x = NULL, y = "%") +
  theme(legend.position = "top")
```
:::
:::::

::: {style="font-size:0.91em; margin-top:4px; padding:5px 12px; background:#fef9e7; border-left:3px solid #c0392b; border-radius:3px;"}
**Interpretation:** χ² (p \< 0.001) confirms bonus composition is not independent of policy type. **COMP_N carries 3.5% Bad-score policyholders**; nearly 3× the rate of other types, a textbook case of adverse selection. **New Business carries 2× the Bad-score rate of Portfolio renewals** (1.6% vs 0.8%), confirming the acquisition channel absorbs materially worse risk before renewal underwriting filters apply.
:::

------------------------------------------------------------------------

## Slide 7 \| Univariate: Geographic & Vehicle Distributions

::::: columns
::: {.column width="34%"}
```{r}
#| label: slide7-geo
#| fig-height: 3.7
#| fig-width: 3.4
geo_df <- insurance_viz |>
  count(municipality_type, circulation_area) |>
  group_by(municipality_type) |>
  mutate(pct = n / sum(n) * 100, n_total = sum(n)) |>
  ungroup()

ggplot(geo_df,
       aes(x = municipality_type, y = pct, fill = circulation_area)) +
  geom_col(position = "stack", width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(round(pct,0),"%\n(n=",comma(n),")")),
            position = position_stack(vjust = 0.5),
            size = 2.3, colour = "white", fontface = "bold") +
  geom_text(data = distinct(geo_df, municipality_type, n_total),
            aes(x = municipality_type, y = 107,
                label = paste0("n=",comma(n_total))),
            inherit.aes = FALSE, size = 2.4, colour = "grey30") +
  scale_fill_manual(values = c("Urban"="#2980b9","Rural"="#8e44ad"), name = NULL) +
  scale_y_continuous(limits = c(0,115), labels = function(x) paste0(x,"%")) +
  labs(title = "Geographic Split",
       x = NULL, y = "%") +
  theme(legend.position = "top")
```
:::

::: {.column width="66%"}
```{r}
#| label: slide7-vehicle-dist
#| fig-height: 3.7
#| fig-width: 6.2
p_vage <- ggplot(insurance_viz |> filter(!is.na(vehicle_age)),
                 aes(x = vehicle_age)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1,
                 fill = "#8e44ad", colour = "white", linewidth = 0.15) +
  geom_density(colour = "#e67e22", linewidth = 0.9) +
  geom_vline(xintercept = median(insurance_viz$vehicle_age, na.rm=TRUE),
             linetype = "dashed", colour = "grey40") +
  annotate("text",
           x = median(insurance_viz$vehicle_age, na.rm=TRUE)+1.5, y=0.022,
           label = paste0("Med ",
             round(median(insurance_viz$vehicle_age, na.rm=TRUE),0),"yr"),
           hjust = 0, size = 2.6, colour = "grey40") +
  labs(title = "Vehicle Age", x = "Years", y = "Density")

p_vval <- ggplot(insurance_viz |> filter(!is.na(vehicle_value)),
                 aes(x = vehicle_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#2980b9", colour = "white", linewidth = 0.15) +
  geom_density(colour = "#c0392b", linewidth = 0.9) +
  scale_x_log10(labels = comma) +
  labs(title = "Vehicle Value (log₁₀)", x = "EUR", y = "Density")

p_vage + p_vval
```
:::
:::::

::: {style="font-size:0.91em; margin-top:4px; padding:5px 12px; background:#eaf4fc; border-left:3px solid #2980b9; border-radius:3px;"}
**Interpretation:** Islands has only \~24,000 policies which makes it too few for statistically reliable loss ratio estimates (wide CIs confirmed in Slide 14). Vehicle age is right-skewed (mode 20–25 yrs), indicating an older fleet where the non-linear severity curve applies at both extremes. Vehicle value is log-normally distributed (median €24,700) making log scale mandatory as linear scale will compress the modal range into an unreadable sliver.
:::

------------------------------------------------------------------------

## Slide 8 \| Univariate: Driver Age Distribution

:::::: columns
::: {.column width="56%"}
```{r}
#| label: slide8-driver-age
#| fig-height: 3.9
#| fig-width: 5.6
med_age <- median(insurance_viz$driver_age, na.rm=TRUE)

ggplot(insurance_viz, aes(x = driver_age)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2,
                 fill = "#27ae60", colour = "white", linewidth = 0.2) +
  geom_density(colour = "#c0392b", linewidth = 1.1) +
  geom_vline(xintercept = c(25, 65), linetype = "dashed",
             colour = "#2c3e50", linewidth = 0.8) +
  annotate("rect", xmin=17, xmax=25, ymin=0, ymax=0.043, alpha=0.08, fill="#e74c3c") +
  annotate("rect", xmin=65, xmax=90, ymin=0, ymax=0.043, alpha=0.08, fill="#e74c3c") +
  annotate("text", x=21, y=0.040, label="Young\nrisk\nzone",
           size=2.6, colour="#c0392b", hjust=0.5) +
  annotate("text", x=73, y=0.040, label="Senior\nrisk\nzone",
           size=2.6, colour="#c0392b", hjust=0.5) +
  annotate("text", x=26, y=0.030, label="25", size=2.8, colour="#2c3e50") +
  annotate("text", x=66, y=0.030, label="65", size=2.8, colour="#2c3e50") +
  geom_vline(xintercept=med_age, linetype="dotted", colour="#2980b9", linewidth=0.8) +
  annotate("text", x=med_age+1.5, y=0.025,
           label=paste0("Median: ",round(med_age,0)),
           size=2.8, colour="#2980b9", hjust=0) +
  scale_x_continuous(breaks=seq(20,90,10)) +
  labs(title="Distribution of Driver Age",
       subtitle="Shaded zones = actuarial high-risk segments (under 25 and over 65)",
       x="Driver Age (years)", y="Density")
```
:::

:::: {.column width="44%"}
::: {style="font-size:0.90em; line-height:1.35;"}
**Portfolio is concentrated in the 35–60 band**.

-   **Under-25 drivers** are markedly under-represented.
-   **Over-65 drivers** taper naturally with licence cessation.

**Statistical implication:** The U-shaped age–risk relationship (young and old drivers claiming more) is well established. However, because both risk extremes are *thinly populated* here, the age signal is statistically diluted; explaining why the Spearman ρ between driver age and claim frequency is near-zero (−0.035) in Slide 12. This is a portfolio composition effect, not evidence that age is irrelevant to risk.
:::
::::
::::::

------------------------------------------------------------------------

## Slide 9 \| Univariate: Summary of Findings

::::::: columns
::::: {.column width="52%"}
::: {style="font-size:0.90em;"}
| Variable | Pattern | Implication |
|:-----------------------|:-----------------------|:-----------------------|
| **Policy type** | COMP_N: 3.5% Bad; TP: 1.6% | Adverse selection in comp. cover |
| **Business type** | NB: 1.6% Bad vs 0.8% Portfolio | Acquisition absorbs worse risk |
| **Municipality** | Islands: n ≈ 24,000 only | Estimates unreliable; need uncertainty buffer |
| **Vehicle age** | Mode 20–25 yrs; right-skewed | Non-linear severity at both extremes |
| **Vehicle value** | Log-normal; median €24,700 | Log-scale severity loadings required |
| **Driver age** | 35–60 dominated; thin extremes | U-shape diluted; signal at extremes |
:::

::: {style="font-size:0.90em; margin-top:6px; padding:5px 10px; background:#eaf4fc; border-left:3px solid #2980b9; border-radius:3px;"}
**Cross-variable insight:** An older vehicle fleet + middle-aged driver base means demographic extremes are bypassed. Risk concentration is instead occurring in the **product and acquisition channel dimensions** (COMP_N adverse selection, NB vs Portfolio gap).
:::
:::::

::: {.column width="48%"}
```{r}
#| label: slide9-nongood-bar
#| fig-height: 3.9
#| fig-width: 4.8
comp_data <- insurance_viz |>
  count(policy_type, bonus_score) |>
  group_by(policy_type) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  filter(bonus_score != "Good") |>
  mutate(policy_type = str_remove_all(policy_type, "\n"))

ggplot(comp_data,
       aes(x = fct_reorder(policy_type, pct, sum),
           y = pct, fill = bonus_score)) +
  geom_col(position = "dodge", width = 0.65) +
  geom_text(aes(label = paste0(round(pct,1),"%")),
            position = position_dodge(width = 0.65),
            hjust = -0.1, size = 2.7) +
  scale_fill_manual(values = c("Neutral"="#e67e22","Bad"="#c0392b"),
                    name = "Bonus Score") +
  scale_y_continuous(limits = c(0,6.5), labels = function(x) paste0(x,"%")) +
  coord_flip() +
  labs(title = "Non-Good Bonus % by Policy Type",
       subtitle = "Higher % = greater adverse selection signal",
       x = NULL, y = "% of policies in type") +
  theme(legend.position = "top")
```
:::
:::::::

------------------------------------------------------------------------

## Slide 10 \| Bivariate: Loss Ratio- Policy Type & Bonus Score

::::: columns
::: {.column width="50%"}
```{r}
#| label: slide10-lr-policy
#| fig-height: 3.4
#| fig-width: 5.0
ggplot(
  insurance_viz |> filter(loss_ratio_capped > 0, total_exposure > 0),
  aes(x = policy_type, y = loss_ratio_capped, fill = policy_type)
) +
  geom_violin(alpha = 0.5, trim = TRUE, show.legend = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA,
               fill = "white", colour = "grey30", show.legend = FALSE) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "#c0392b") +
  annotate("text", x = 0.65, y = 1.08, label = "Break-even",
           size = 2.6, colour = "#c0392b", hjust = 0) +
  scale_fill_viridis_d(option = "D") +
  scale_x_discrete(labels = function(x) str_wrap(x, 11)) +
  stat_compare_means(method = "kruskal.test", label.x = 1.2,
                     label.y = quantile(insurance_viz$loss_ratio_capped,
                                        0.995, na.rm=TRUE) * 0.95,
                     size = 2.6) +
  labs(title = "Loss Ratio by Policy Type",
       subtitle = "Kruskal-Wallis p < 0.05",
       x = NULL, y = "Loss Ratio")
```
:::

::: {.column width="50%"}
```{r}
#| label: slide10-lr-bonus
#| fig-height: 3.4
#| fig-width: 5.0
ggplot(
  insurance_viz |> filter(loss_ratio_capped > 0, total_premium > 0),
  aes(x = bonus_score, y = loss_ratio_capped, fill = bonus_score)
) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.6) +
  geom_boxplot(width = 0.14, outlier.shape = NA,
               fill = "white", colour = "grey30") +
  stat_compare_means(
    comparisons = list(c("Good","Neutral"),c("Neutral","Bad"),c("Good","Bad")),
    method = "wilcox.test", label = "p.signif", size = 2.8) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "#c0392b") +
  scale_fill_manual(values = bonus_pal) +
  labs(title = "Loss Ratio by Bonus Score",
       subtitle = "All pairwise Wilcoxon p < 0.05",
       x = NULL, y = "Loss Ratio") +
  theme(legend.position = "none")
```
:::
:::::

::: {style="font-size:0.91em; margin-top:4px; padding:5px 12px; background:#fef9e7; border-left:3px solid #c0392b; border-radius:3px;"}
**Interpretation:** COMP_N has the **highest median loss ratio and widest violin**- simultaneously the least profitable and most volatile product. Violin width reveals substantial mass above break-even, not just outliers (Kruskal-Wallis p \< 0.05). Loss ratio increases **monotonically Good→Bad** with all pairwise tests significant. Bad-score half-eye shows a heavier right tail- bad policyholders drive disproportionate *large-loss* exposure, which is what drives capital requirements, not the mean.
:::

------------------------------------------------------------------------

## Slide 11 \| Bivariate: Claim Severity - Fuel Type & Vehicle Age

::::: columns
::: {.column width="42%"}
```{r}
#| label: slide11-sev-fuel
#| fig-height: 3.7
#| fig-width: 4.2
ggplot(
  insurance_viz |>
    filter(!is.na(severity_capped), !is.na(fuel_type), severity_capped > 0),
  aes(x = fuel_type, y = severity_capped, fill = fuel_type)
) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.65) +
  geom_boxplot(width = 0.14, outlier.shape = NA,
               fill = "white", colour = "grey30") +
  stat_compare_means(method = "wilcox.test", label = "p.format",
                     label.x = 1.5, label.y.npc = 0.88, size = 3) +
  scale_y_log10(labels = comma,
                breaks = c(10,50,100,500,1000,5000), limits = c(10,7000)) +
  scale_fill_manual(values = c("Diesel"="#2980b9","Gasoline"="#e67e22")) +
  labs(title = "Severity by Fuel Type",
       subtitle = "Log scale | zero-incurred excluded",
       x = NULL, y = "Severity (EUR, log)") +
  theme(legend.position = "none")
```
:::

::: {.column width="58%"}
```{r}
#| label: slide11-sev-vage
#| fig-height: 3.7
#| fig-width: 5.6
ggplot(
  insurance_viz |>
    filter(!is.na(vehicle_age), !is.na(severity_capped), severity_capped > 0),
  aes(x = vehicle_age, y = severity_capped)
) +
  geom_hex(bins = 50, aes(fill = after_stat(log(count)))) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
              colour = "#c0392b", linewidth = 1, se = TRUE) +
  scale_fill_viridis_c(option = "magma", name = "log(n)") +
  scale_y_log10(labels = comma,
                breaks = c(10,100,1000,5000), limits = c(5,7500)) +
  labs(title = "Severity vs Vehicle Age",
       subtitle = "Hexbin + GAM smoother | log y-axis",
       x = "Vehicle Age (years)", y = "Severity (EUR, log)")
```
:::
:::::

::: {style="font-size:0.91em; margin-top:4px; padding:5px 12px; background:#eaf4fc; border-left:3px solid #2980b9; border-radius:3px;"}
**Interpretation:** Diesel vehicles show **statistically higher severity** (Wilcoxon p = 0.04) - a consistent median *location shift*, not a tail effect, supporting a 5–10% diesel severity loading. Vehicle age shows a **non-linear U-shape**: severity peaks for new vehicles (high replacement value), declines through mid-years, then rises again for very old vehicles (rare parts, specialist labour). A simple declining age factor in the rating engine is therefore incorrect.
:::

------------------------------------------------------------------------

## Slide 12 \| Bivariate: Demographic Drivers

::::: columns
::: {.column width="54%"}
```{r}
#| label: slide12-ridge-age
#| fig-height: 3.7
#| fig-width: 5.4
ggplot(
  insurance_viz |>
    filter(loss_ratio_capped > 0, total_exposure > 0, !is.na(driver_age_band)),
  aes(x = loss_ratio_capped, y = driver_age_band,
      fill = factor(after_stat(quantile)))
) +
  stat_density_ridges(
    geom = "density_ridges_gradient", calc_ecdf = TRUE,
    quantiles = 4, quantile_lines = TRUE,
    scale = 1.9, rel_min_height = 0.005, bandwidth = 0.045) +
  scale_fill_viridis_d(name = "Quartile", option = "D", alpha = 0.85) +
  scale_x_continuous(limits = c(0,5), expand = c(0,0)) +
  scale_y_discrete(expand = expansion(add = c(0.3, 2.0))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  labs(title = "Loss Ratio by Driver Age Band",
       subtitle = "Ridgeline + quartile shading | dashed = break-even",
       x = "Loss Ratio", y = NULL) +
  theme_ridges(font_size = 10) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
        legend.position = "right", legend.text = element_text(size = 8))
```
:::

::: {.column width="46%"}
```{r}
#| label: slide12-heatmap
#| fig-height: 3.7
#| fig-width: 4.4
hm_data <- insurance_viz |>
  filter(!is.na(driver_age_band), !is.na(vehicle_age_band), total_exposure > 0) |>
  group_by(driver_age_band, vehicle_age_band) |>
  summarise(mean_freq = mean(claim_frequency, na.rm=TRUE), n=n(), .groups="drop") |>
  filter(n >= 30) |>
  mutate(mean_freq = pmin(mean_freq, 2.0))

mid_val <- median(hm_data$mean_freq, na.rm=TRUE)

ggplot(hm_data,
       aes(x = vehicle_age_band, y = driver_age_band, fill = mean_freq)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(mean_freq,2),
                colour = mean_freq > mid_val),
            size = 2.4, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = c("TRUE"="white","FALSE"="grey20")) +
  scale_fill_gradient2(low="#2166ac", mid="#f7f7f7", high="#d6604d",
                       midpoint = mid_val, name = "Mean\nFreq") +
  labs(title = "Claim Freq: Age Interaction",
       subtitle = "Driver × Vehicle Age | mean; n<30 excluded",
       x = "Vehicle Age Band", y = NULL) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
```
:::
:::::

::: {style="font-size:0.91em; margin-top:4px; padding:5px 12px; background:#eaf4fc; border-left:3px solid #2980b9; border-radius:3px;"}
**Interpretation:** Ridgelines confirm the **actuarial U-shape** - 18–25 band widest and most right-skewed; 36–55 most compact and closest to break-even; 65+ moderately elevated. Heatmap reveals an interaction: **65+ drivers in 0–3-year vehicles show elevated claim frequency**; consistent with older drivers adapting to unfamiliar new vehicle technology. Driver age and years_licensed both show near-zero Spearman ρ, a consequence of the under-25 and over-65 extremes being thinly represented in this portfolio (confirmed in Slide 8).
:::

------------------------------------------------------------------------

## Slide 13 \| Bivariate: Brand, Status & Behavioural Drivers

::::: columns
::: {.column width="46%"}
```{r}
#| label: slide13-brand-panels
#| fig-height: 3.7
#| fig-width: 4.6
brand_data <- insurance_viz |> filter(!is.na(brand_tier), total_exposure > 0)

p_freq_b <- ggplot(brand_data, aes(x = brand_tier, y = claim_frequency)) +
  stat_pointinterval(.width = c(0.95,0.99), point_size = 2,
                     aes(colour = factor(after_stat(.width)))) +
  scale_colour_manual(name="CI",
    values=c("0.95"="#2980b9","0.99"="#e67e22"), labels=c("95%","99%")) +
  scale_x_discrete(labels = function(x) str_wrap(x, 9)) +
  labs(title = "Claim Frequency", x = NULL, y = "Freq") +
  theme(legend.position = "top")

p_sev_b <- ggplot(
  brand_data |> filter(!is.na(severity_capped), severity_capped > 0),
  aes(x = brand_tier, y = severity_capped)) +
  stat_pointinterval(.width = c(0.95,0.99), point_size = 2,
                     aes(colour = factor(after_stat(.width)))) +
  scale_colour_manual(name="CI",
    values=c("0.95"="#2980b9","0.99"="#e67e22"), labels=c("95%","99%")) +
  scale_y_log10(labels=comma,
                breaks=c(10,100,1000,5000), limits=c(10,7500)) +
  scale_x_discrete(labels = function(x) str_wrap(x, 9)) +
  labs(title = "Claim Severity (log)", x = NULL, y = "EUR (log)") +
  theme(legend.position = "top")

p_freq_b / p_sev_b
```
:::

::: {.column width="54%"}
```{r}
#| label: slide13-glass
#| fig-height: 3.7
#| fig-width: 5.2
glass_data <- insurance_viz |>
  filter(total_exposure > 0, claim_frequency < 8) |>
  mutate(glass_label = if_else(had_glass_claim,"Filed glass claim","No glass claim"))

ggplot(glass_data,
       aes(x = claim_frequency, fill = glass_label, colour = glass_label)) +
  geom_density(alpha = 0.38, linewidth = 0.9) +
  scale_fill_manual(
    values = c("Filed glass claim"="#c0392b","No glass claim"="#2980b9"), name=NULL) +
  scale_colour_manual(
    values = c("Filed glass claim"="#922b21","No glass claim"="#1a5276"), name=NULL) +
  scale_x_continuous(limits = c(0,5)) +
  annotate("text", x=1.5, y=1.5,
           label="Glass claimers:\nbimodal at 1.0 & 2.0",
           size=2.8, colour="#922b21", hjust=0) +
  annotate("segment", x=1.4, xend=1.1, y=1.48, yend=1.82,
           colour="#922b21", linewidth=0.5,
           arrow=arrow(length=unit(0.1,"cm"))) +
  labs(title = "Glass Claimers vs Non-Claimers",
       subtitle = "Gateway claim confirmed: bimodal distribution",
       x = "Claim Frequency (events/policy year)", y = "Density") +
  theme(legend.position = "top")
```
:::
:::::

::: {style="font-size:0.91em; margin-top:4px; padding:5px 12px; background:#fef9e7; border-left:3px solid #c0392b; border-radius:3px;"}
**Interpretation:** Brand tier does **not** discriminate claim frequency (all tiers identical), but Luxury's 99% CI extends further upward for severity, confirming disproportionately costly extreme claims. Apply brand tier as a **severity tail loading**, not a frequency loading. Glass claimers show a **bimodal distribution peaking at 1.0 and 2.0 events/year**, they almost never appear at zero frequency. Glass claims are confirmed gateway events that co-occur with broader claims activity; a 10–15% frequency loading at renewal for prior-year glass claimers is supported.
:::

------------------------------------------------------------------------

## Slide 14 \| Bivariate: Loyalty & Geographic Drivers

::::: columns
::: {.column width="35%"}
```{r}
#| label: slide14-cancel
#| fig-height: 3.7
#| fig-width: 3.4
cancel_df <- insurance_viz |>
  count(bonus_score, policy_status) |>
  group_by(bonus_score) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()

ggplot(cancel_df,
       aes(x = bonus_score, y = pct, fill = policy_status)) +
  geom_col(position = "stack", width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(round(pct,0),"%")),
            position = position_stack(vjust = 0.5),
            size = 3, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Active"="#27ae60","Cancelled"="#c0392b"), name = "Status") +
  scale_y_continuous(labels = function(x) paste0(x,"%")) +
  labs(title = "Cancellation\nby Bonus Score",
       subtitle = "Bad: 75% stay\nGood: 90% stay",
       x = "Bonus Score", y = "%") +
  theme(legend.position = "top")
```
:::

::: {.column width="65%"}
```{r}
#| label: slide14-geo-ci
#| fig-height: 3.7
#| fig-width: 6.2
geo_ci <- insurance_viz |>
  filter(loss_ratio_capped > 0, total_exposure > 0) |>
  mutate(
    geo_seg = paste0(municipality_type, "\n(", circulation_area, ")"),
    geo_seg = fct_reorder(geo_seg, loss_ratio_capped, median)
  )

ggplot(geo_ci, aes(x = geo_seg, y = loss_ratio_capped)) +
  stat_pointinterval(.width = c(0.95,0.99), point_size = 2.8,
                     aes(colour = factor(after_stat(.width)))) +
  scale_colour_manual(name = "CI",
    values = c("0.95"="#2980b9","0.99"="#e67e22"),
    labels = c("95% CI","99% CI")) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  annotate("text", x=0.65, y=1.08, label="Break-even",
           hjust=0, size=2.8, colour="grey40") +
  labs(title = "Loss Ratio by Geographic Segment",
       subtitle = "Wide island CIs = pricing uncertainty",
       x = NULL, y = "Loss Ratio") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 8.5))
```
:::
:::::

::: {style="font-size:0.91em; margin-top:4px; padding:5px 12px; background:#fef9e7; border-left:3px solid #c0392b; border-radius:3px;"}
**Interpretation:** **Bad-score policyholders cancel at 25% but 75% stay**; the stickiest cohort, likely lacking competitive alternatives. The insurer's attrition process fails to shed its worst risks; actively pricing Bad-score COMP_N to not-renew is more effective than passive stickiness. **Urban segments show consistently higher loss ratios than rural**. Island segments have dramatically wide CIs; the observed median loss ratio is statistically unreliable; pricing should apply an explicit uncertainty buffer at the 95th percentile of the CI, not the point estimate.
:::

------------------------------------------------------------------------

## Slide 15 \| Conclusion & Reflection

::::::: columns
:::: {.column width="52%"}
::: {style="font-size:0.90em; line-height:1.35;"}
**What this exercise demonstrates**

Visual analytics combined with statistical rigour, provides structured, evidence-based insights. Distributional visualisation, non-parametric testing, and explicit uncertainty quantification reveal patterns that summary statistics alone miss.

**The role of the data analyst**

A data analyst **surfaces patterns and quantifies associations; they do not make business decisions.** Translating findings into business strategies requires actuarial judgement, regulatory context, and commercial considerations beyond analytical scope.

**Crucially: correlation ≠ causation** e.g., glass claimers showing higher claim frequencies describes a *pattern* but it does not establish that glass claims *cause* subsequent claims. Establishing causality requires controlled study designs that cross-sectional analysis cannot provide.

**Limitations**

-   `years_licensed` mis-specified; re-derivation needed

-   Cross-sectional data; no causal inference
:::
::::

:::: {.column width="48%"}
::: {style="font-size:0.90em; line-height:1.35;"}
**Further work required**

🔬 Re-derive `years_licensed` as genuine experience years and retest as risk predictor

🔬 Investigate interaction terms (COMP_N × Bad bonus; Luxury × Neutral) via multivariate GLM

🔬 Temporal loss ratio trend analysis across 2022–2024 to detect deteriorating segments

🔬 Apply credibility weighting to thin-data cells before pricing decisions

📊 Enrich with geospatial data to model risk at postcode level

📊 Link multi-year records to confirm glass claim predictive value in Year 2

⚙️ Validate all findings with claims and pricing teams before rating engine changes

⚙️ Commission actuarial review of COMP_N pricing adequacy as an **immediate priority**
:::
::::
:::::::
